---
title: "Constrained ordination (db-RDA) on simulated spesim data"
output:
  rmarkdown::html_vignette:
    toc: true
    highlight: null
vignette: >
  %\VignetteIndexEntry{Constrained ordination (db-RDA) on simulated spesim data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>",
  warning = FALSE,
  message = FALSE,
  fig.width = 4.5,
  fig.height = 3.0,
  out.width = "85%",
  dpi = 300
)

ggplot2::theme_set(ggplot2::theme_bw(base_size = 8))
```

This vignette shows how to use **spesim** to generate a realistic (but known)
community dataset for practising constrained ordination.

We will:

1. Simulate a community in an irregular domain with **strong environmental gradients**.
2. Add **positive and negative interspecific interactions** (local neighbourhood effects).
3. Compare two **transect sampling designs**:
   - one transect aligned **along** the temperature gradient
   - one transect set **across** (perpendicular to) the temperature gradient
4. For each design, obtain:
   - a **site × species** abundance table
   - a **site × environment** table (mean environment per quadrat)
5. Fit **distance-based RDA (db-RDA)** models using **vegan**’s `capscale()`.
6. Arrange paired figures (`a`, `b`) and summarise key results in a **gt** table.

## 1) Libraries

```{r}
library(spesim)
library(dplyr)
library(ggplot2)
library(vegan)
library(patchwork)

has_gt <- requireNamespace("gt", quietly = TRUE)
```

## 2) Configure a simulation with strong gradients + interactions

We start from the packaged basic init file and then modify it.

```{r}
`%||%` <- function(a, b) if (!is.null(a)) a else b

init_file <- system.file("examples/spesim_init_basic.txt", package = "spesim")
P <- load_config(init_file)

# --- keep the vignette fast-ish
P$N_INDIVIDUALS <- 3000L
P$N_SPECIES <- 16L
P$N_QUADRATS <- 10L
P$SAMPLING_RESOLUTION <- 70L
P$ENVIRONMENTAL_NOISE <- 0.02
P$ADVANCED_ANALYSIS <- FALSE
quadrat_size_scale <- 0.8  # reduce quadrat side lengths by 20%

# --- strong filtering (narrow tolerances)
# A-J respond to temperature, K-L to elevation, M-P have no gradient response.
P$GRADIENT_SPECIES <- LETTERS[1:12]
P$GRADIENT_ASSIGNMENTS <- c(rep("temperature", 10), rep("elevation", 2))
P$GRADIENT_OPTIMA <- c(
  stats::setNames(seq(0.10, 0.90, length.out = 10), LETTERS[1:10]),
  K = 0.30,
  L = 0.75
)
P$GRADIENT_TOLERANCE <- 0.06

# --- local interactions
# Values < 1 weaken neighbours (negative interactions); > 1 strengthen (positive).
P$INTERACTION_RADIUS <- 1.0
P$INTERACTIONS_EDGELIST <- c(
  # negative (competition/inhibition)
  "B,A,0.80",
  "C,A,0.85",
  "D,B,0.75",
  "E,D,0.80",

  # mild background pressure for some focal species (from all neighbours)
  "B,*,0.98",
  "C,*,0.98",

  # positive (facilitation)
  "F,G,1.25",
  "G,F,1.15",
  "H,I,1.20",
  "I,H,1.10"
)

# Reproducibility
P$SEED <- 2026L
```

## 3) Two sampling designs and shared analysis helpers

To isolate the effect of sampling design, we keep the same underlying simulated
community and only change transect orientation.

```{r}
# Build a single deterministic domain that both designs will share.
set.seed(P$SEED)
domain <- create_sampling_domain()

# Temperature is generated from normalised coordinates:
#   temp ~ 0.7*x_norm + 0.3*y_norm
# where x_norm = (x - xmin)/(xmax-xmin), so the true gradient direction in
# x–y space depends on the domain bounding box aspect ratio.
bb <- sf::st_bbox(domain)
xrange <- as.numeric(bb["xmax"] - bb["xmin"])
yrange <- as.numeric(bb["ymax"] - bb["ymin"])

# Bearing of increasing temperature (0 = North, 90 = East)
temp_angle <- (atan2(0.7 / xrange, 0.3 / yrange) * 180 / pi) %% 360

# Perpendicular (approx. along isotherms; minimal temperature change)
across_angle <- (temp_angle + 90) %% 360

design_angles <- tibble::tibble(
  design = c("Along temperature gradient", "Across temperature gradient"),
  transect_angle_deg = c(temp_angle, across_angle)
)
design_angles
```

```{r}
safe_cor <- function(x, y) {
  ok <- is.finite(x) & is.finite(y)
  if (sum(ok) < 3) return(NA_real_)
  if (stats::sd(x[ok]) == 0 || stats::sd(y[ok]) == 0) return(NA_real_)
  stats::cor(x[ok], y[ok])
}

build_sampling_plot <- function(res, title, subtitle = NULL) {
  ctr <- sf::st_coordinates(sf::st_centroid(res$quadrats))
  tran <- data.frame(
    quadrat_id = res$quadrats$quadrat_id,
    x = ctr[, 1],
    y = ctr[, 2]
  ) |>
    dplyr::arrange(quadrat_id)

  ggplot() +
    geom_raster(
      data = res$env_gradients,
      aes(x = x, y = y, fill = temperature_C),
      alpha = 0.9,
      interpolate = TRUE
    ) +
    viridis::scale_fill_viridis(name = "Temperature (deg C)", option = "mako") +
    geom_sf(data = res$domain, fill = NA, colour = "grey20", linewidth = 0.4) +
    geom_path(
      data = tran,
      aes(x = x, y = y),
      linewidth = 0.45,
      colour = "white",
      alpha = 0.8
    ) +
    geom_sf(data = res$quadrats, fill = NA, colour = "black", linewidth = 0.45) +
    coord_sf(expand = FALSE) +
    theme_bw(base_size = 8) +
    labs(title = title, subtitle = subtitle, x = NULL, y = NULL)
}

build_cap_plot <- function(sc_sites, sc_bp, title) {
  p <- ggplot(sc_sites, aes(x = CAP1, y = CAP2)) +
    geom_hline(yintercept = 0, linewidth = 0.2, colour = "grey70") +
    geom_vline(xintercept = 0, linewidth = 0.2, colour = "grey70") +
    geom_point(aes(colour = temp_bin), size = 1.6, alpha = 0.85) +
    coord_equal() +
    scale_colour_brewer(palette = "RdYlBu", direction = -1, name = "Temp quartile") +
    labs(
      x = "db-RDA axis 1 (CAP1)",
      y = "db-RDA axis 2 (CAP2)",
      title = title
    )

  if (nrow(sc_bp) > 0) {
    p <- p +
      geom_segment(
        data = sc_bp,
        aes(x = 0, y = 0, xend = CAP1, yend = CAP2),
        inherit.aes = FALSE,
        arrow = arrow(length = unit(0.14, "cm")),
        linewidth = 0.3,
        colour = "black"
      ) +
      geom_text(
        data = sc_bp,
        aes(x = CAP1, y = CAP2, label = var),
        inherit.aes = FALSE,
        size = 2.5,
        vjust = -0.6
      )
  }

  p
}

analyse_design <- function(P_base, domain, label, angle_deg) {
  P2 <- P_base
  P2$SAMPLING_SCHEME <- "transect"
  P2$N_TRANSECTS <- 1L
  P2$N_QUADRATS_PER_TRANSECT <- P2$N_QUADRATS
  P2$TRANSECT_ANGLE <- angle_deg

  # Build deterministic species/environment first, then re-place
  # reduced-size transect quadrats for this comparison.
  res <- spesim_run(P2, domain = domain, write_outputs = FALSE, seed = P2$SEED, quiet = TRUE)

  quadrat_size_reduced <- as.numeric(res$P$QUADRAT_SIZE) * quadrat_size_scale
  quadrats <- place_quadrats_transect(
    domain = res$domain,
    n_transects = P2$N_TRANSECTS,
    n_quadrats_per_transect = P2$N_QUADRATS_PER_TRANSECT,
    quadrat_size = quadrat_size_reduced,
    angle = angle_deg
  )
  if (nrow(quadrats) == 0) {
    stop("Failed to place reduced-size quadrats.")
  }
  res$quadrats <- quadrats

  all_species <- LETTERS[1:P2$N_SPECIES]
  A <- create_abundance_matrix(res$species_dist, res$quadrats, all_species)
  E <- calculate_quadrat_environment(res$env_gradients, res$quadrats, sf::st_crs(res$domain))
  dat <- dplyr::left_join(E, A, by = "site")
  stopifnot(nrow(dat) == nrow(A))

  Y <- dat |> dplyr::select(-site, -temperature_C, -elevation_m, -rainfall_mm)
  X <- dat |> dplyr::select(site, temperature_C, elevation_m, rainfall_mm)
  Xz <- vegan::decostand(dplyr::select(X, -site), method = "standardize")

  cap_model <- vegan::capscale(
    Y ~ temperature_C + elevation_m + rainfall_mm,
    data = Xz,
    distance = "bray",
    add = TRUE
  )

  set.seed(P2$SEED)
  model_test <- vegan::anova.cca(cap_model, permutations = 199)
  model_p <- as.numeric(model_test$`Pr(>F)`[1])
  adj_r2 <- as.numeric(vegan::RsquareAdj(cap_model)$adj.r.squared)
  vif_vals <- tryCatch(vegan::vif.cca(cap_model), error = function(e) NA_real_)
  vif_vals <- as.numeric(vif_vals)
  vif_vals <- vif_vals[is.finite(vif_vals)]
  vif_max <- if (length(vif_vals)) max(vif_vals) else NA_real_

  sc_sites <- as.data.frame(vegan::scores(cap_model, display = "sites", scaling = 2))
  sc_bp <- as.data.frame(vegan::scores(cap_model, display = "bp", scaling = 2))
  sc_sites$site <- dat$site
  sc_sites <- dplyr::left_join(sc_sites, X, by = "site")
  sc_sites$temp_bin <- factor(dplyr::ntile(sc_sites$temperature_C, 4), labels = paste0("Q", 1:4))

  arrow_mul <- 1.3
  sc_bp$var <- rownames(sc_bp)
  if (nrow(sc_bp) > 0) {
    sc_bp$CAP1 <- sc_bp$CAP1 * arrow_mul
    sc_bp$CAP2 <- sc_bp$CAP2 * arrow_mul
  }

  cor_tbl <- sc_sites |>
    dplyr::summarise(
      cor_cap1_temp = safe_cor(CAP1, temperature_C),
      cor_cap1_elev = safe_cor(CAP1, elevation_m),
      cor_cap1_rain = safe_cor(CAP1, rainfall_mm)
    ) |>
    dplyr::mutate(design = label, .before = 1)

  metrics <- cor_tbl |>
    dplyr::mutate(
      transect_angle_deg = angle_deg,
      n_quadrats = nrow(res$quadrats),
      temp_range_C = diff(range(sc_sites$temperature_C, na.rm = TRUE)),
      adj_r2 = adj_r2,
      model_p = model_p,
      max_vif = vif_max,
      .before = cor_cap1_temp
    ) |>
    dplyr::select(
      design,
      transect_angle_deg,
      n_quadrats,
      temp_range_C,
      adj_r2,
      model_p,
      cor_cap1_temp,
      cor_cap1_elev,
      cor_cap1_rain,
      max_vif
    )

  sampling_plot <- build_sampling_plot(
    res,
    title = label,
    subtitle = sprintf("angle = %.1f deg", angle_deg)
  )
  cap_plot <- build_cap_plot(sc_sites, sc_bp, title = label)

  list(
    label = label,
    angle = angle_deg,
    res = res,
    A = A,
    E = E,
    dat = dat,
    cap = cap_model,
    cor_tbl = cor_tbl,
    metrics = metrics,
    sampling_plot = sampling_plot,
    cap_plot = cap_plot
  )
}
```

## 4) Run both designs

```{r}
# 'domain' was created above so that both designs share the same geometry.
along <- analyse_design(
  P_base = P,
  domain = domain,
  label = "Along temperature gradient",
  angle_deg = temp_angle
)
across <- analyse_design(
  P_base = P,
  domain = domain,
  label = "Across temperature gradient",
  angle_deg = across_angle
)
```

Quick check that only sampling differs (same simulated species cloud):

```{r}
isTRUE(all.equal(
  sf::st_coordinates(along$res$species_dist),
  sf::st_coordinates(across$res$species_dist)
))
```

## 5) Figure 1: sampling designs (`a`, `b`)

```{r fig-sampling, fig.width=7.2, fig.height=3.4, out.width="100%", fig.cap="Figure 1. Quadrat placement over the temperature surface: (a) along-gradient transect, (b) across-gradient transect."}
fig_sampling <- patchwork::wrap_plots(
  along$sampling_plot,
  across$sampling_plot,
  ncol = 2,
  guides = "collect"
) +
  patchwork::plot_annotation(tag_levels = "a")

fig_sampling
```

## 6) Full db-RDA workflow for each design

### 6.1 Site × species and site × environment tables

```{r}
dplyr::bind_rows(
  tibble::tibble(design = along$label, n_sites = nrow(along$A), n_species_cols = ncol(along$A) - 1L),
  tibble::tibble(design = across$label, n_sites = nrow(across$A), n_species_cols = ncol(across$A) - 1L)
)
```

### 6.2 db-RDA model summaries

```{r}
along$cap
across$cap
```

## 7) Figure 2: db-RDA ordinations (`a`, `b`)

```{r fig-ordination, fig.width=7.2, fig.height=3.4, out.width="100%", fig.cap="Figure 2. db-RDA ordination biplots for each sampling design."}
fig_ordination <- patchwork::wrap_plots(
  along$cap_plot,
  across$cap_plot,
  ncol = 2,
  guides = "collect"
) +
  patchwork::plot_annotation(tag_levels = "a")

fig_ordination
```

## 8) Figure 3: sanity check (CAP1 vs temperature; `a`, `b`)

```{r fig-sanity, fig.width=7.2, fig.height=3.2, out.width="100%", fig.cap="Figure 3. CAP1 tracks the sampled temperature range differently under the two designs."}
p_sanity_along <- ggplot(along$dat, aes(x = temperature_C, y = vegan::scores(along$cap, display = "sites", scaling = 2)[, "CAP1"])) +
  geom_point(size = 1.2, alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, colour = "black") +
  theme_bw(base_size = 8) +
  labs(x = "Temperature (deg C)", y = "CAP1", title = along$label)

p_sanity_across <- ggplot(across$dat, aes(x = temperature_C, y = vegan::scores(across$cap, display = "sites", scaling = 2)[, "CAP1"])) +
  geom_point(size = 1.2, alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, colour = "black") +
  theme_bw(base_size = 8) +
  labs(x = "Temperature (deg C)", y = "CAP1", title = across$label)

patchwork::wrap_plots(p_sanity_along, p_sanity_across, ncol = 2) +
  patchwork::plot_annotation(tag_levels = "a")
```

## 9) Key results summary table (`gt`)

```{r}
summary_tbl <- dplyr::bind_rows(along$metrics, across$metrics)

if (has_gt) {
  summary_tbl |>
    gt::gt() |>
    gt::fmt_number(
      columns = c(transect_angle_deg, temp_range_C, adj_r2, cor_cap1_temp, cor_cap1_elev, cor_cap1_rain, max_vif),
      decimals = 3
    ) |>
    gt::fmt_number(columns = model_p, decimals = 4) |>
    gt::cols_label(
      design = "Sampling design",
      transect_angle_deg = "Transect angle (deg)",
      n_quadrats = "Quadrats",
      temp_range_C = "Sampled temp range (deg C)",
      adj_r2 = "Adj. R^2",
      model_p = "Model p-value",
      cor_cap1_temp = "cor(CAP1, temp)",
      cor_cap1_elev = "cor(CAP1, elev)",
      cor_cap1_rain = "cor(CAP1, rain)",
      max_vif = "Max VIF"
    ) |>
    gt::tab_header(
      title = "db-RDA comparison by sampling design",
      subtitle = "Along-gradient vs across-gradient transects"
    )
} else {
  knitr::kable(summary_tbl, digits = 3)
}
```

## 10) Why this is useful for teaching

Because spesim generates:
- the **site × species** matrix (`abund_matrix`), and
- the corresponding **site × environment** matrix (via `calculate_quadrat_environment()`),

you can repeatedly create controlled datasets where you *know the truth* (strong
gradients and interactions were imposed) and then practise ordination workflows
that aim to recover the dominant drivers under alternative sampling designs.

### Suggested exercises

- Increase/decrease `GRADIENT_TOLERANCE` and see how the constrained R² changes.
- Turn interactions off (`INTERACTION_RADIUS <- 0`) and compare db-RDA fits.
- Compare single-transect, multi-transect, and non-transect designs to test
  sensitivity of constrained ordination outcomes to sampling geometry.
