---
title: "Quadrat placement schemes in spesim"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Quadrat placement schemes in spesim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

This vignette shows the **five quadrat placement schemes** available in **spesim**, with reproducible code and simple visualisations for each:

- `random` → [`place_quadrats()`]
- `tiled` → [`place_quadrats_tiled()`]
- `systematic` → [`place_quadrats_systematic()`]
- `transect` → [`place_quadrats_transect()`]
- `voronoi` → [`place_quadrats_voronoi()`]

All schemes place **axis-aligned rectangular quadrats** (i.e., not rotated). Each function attempts to ensure quadrats lie fully within the sampling domain; specific constraints differ by scheme.

## Setup: a shared irregular domain

```{r}
library(spesim)
library(sf)
library(ggplot2)

set.seed(1)
domain <- create_sampling_domain()

# Quadrat size is given in the domain's coordinate units.
# The default domain is synthetic and uses an arbitrary planar scale.
quadrat_size <- c(1.5, 1.5)
```

A small helper to plot the domain + quadrats consistently:

```{r}
plot_quadrats <- function(domain, quadrats, title) {
  ggplot() +
    geom_sf(data = domain, fill = NA, colour = "#22223b", linewidth = 0.7) +
    geom_sf(data = quadrats, fill = "#4a4e69", alpha = 0.15, colour = "#22223b", linewidth = 0.5) +
    geom_sf_text(
      data = suppressWarnings(sf::st_centroid(quadrats)),
      aes(label = quadrat_id),
      size = 3,
      colour = "#22223b"
    ) +
    coord_sf(datum = NA) +
    labs(title = title) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
}
```

## 1) Random (non-overlapping) placement

**Use when:** you want *simple random sampling* and you don’t want quadrats overlapping.

**Mechanics:** rejection sampling of random centres (uniform over a shrunk bounding box), accepting only quadrats that are **fully inside** the domain and **do not overlap** already accepted quadrats.

```{r}
set.seed(2)
q_random <- place_quadrats(domain, n_quadrats = 20, quadrat_size = quadrat_size)
plot_quadrats(domain, q_random, "Random placement (non-overlapping)")
```

**Notes:**

- If the domain is small or the quadrats are large, you may get fewer quadrats than requested (with a warning).
- This is often a good default for teaching.

## 2) Tiled placement

**Use when:** you want quadrats drawn from a **regular tiling**, with strict non-overlap.

**Mechanics:** a regular grid of cells (`st_make_grid`) at the quadrat size is created; tiles fully inside the domain are candidates; `n_quadrats` are sampled from these candidates.

```{r}
set.seed(3)
q_tiled <- place_quadrats_tiled(domain, n_quadrats = 20, quadrat_size = quadrat_size)
plot_quadrats(domain, q_tiled, "Tiled placement (grid cells inside domain)")
```

**Notes:**

- Returned quadrats never overlap.
- If the domain has narrow "necks", many grid cells may be excluded.

## 3) Systematic placement

**Use when:** you want a **regular lattice** of quadrats (systematic sampling).

**Mechanics:** centres are placed on an `nx × ny` grid spanning the domain bounding box; quadrats centred on those points are retained only if they fit entirely in the domain.

```{r}
set.seed(4)
q_systematic <- place_quadrats_systematic(domain, n_quadrats = 25, quadrat_size = quadrat_size)
plot_quadrats(domain, q_systematic, "Systematic placement (regular lattice)")
```

**Notes:**

- The *actual* number returned may differ from `n_quadrats` because the grid is filtered by the domain boundary.
- Quadrats may overlap if the grid spacing is smaller than the quadrat size (this depends on the inferred grid density).

## 4) Transect placement

**Use when:** you want **transect-based sampling** (parallel lines across the domain) with evenly spaced quadrats along each transect.

**Mechanics:** `n_transects` parallel lines are created across a *safe area* (domain shrunk by half the quadrat diagonal). Along each transect segment, `n_quadrats_per_transect` quadrats are placed.

```{r}
set.seed(5)
q_transect <- place_quadrats_transect(
  domain,
  n_transects = 2,
  n_quadrats_per_transect = 8,
  quadrat_size = quadrat_size,
  angle = 60
)
plot_quadrats(domain, q_transect, "Transect placement (2 transects, angle = 60°)")
```

**Notes:**

- The quadrats themselves remain axis-aligned, even if transects run at an angle.
- If the safe area becomes too small (large quadrats, narrow domain), few or no quadrats may be returned.

## 5) Voronoi-based placement

**Use when:** you want sampling locations that are **well-spaced** in complex domains.

**Mechanics:** a dense set of random seed points are placed in the domain; a Voronoi tessellation is computed; cells are screened for whether they can contain the quadrat (based on an inscribed circle heuristic). Quadrats are placed at cell centres.

```{r}
set.seed(6)
# Note: very large voronoi_seed_factor values tend to create many small cells,
# which can make it impossible to fit larger quadrats. If you get an empty
# result, either reduce `voronoi_seed_factor` or use a smaller quadrat_size.
q_voronoi <- place_quadrats_voronoi(
  domain,
  n_quadrats = 20,
  quadrat_size = c(1.2, 1.2),
  voronoi_seed_factor = 4
)
plot_quadrats(domain, q_voronoi, "Voronoi placement (seed_factor = 4; size = 1.2×1.2)")
```

**Notes:**

- For narrow or very irregular domains, increase `voronoi_seed_factor` (at a computational cost).
- The current implementation focuses on finding feasible, nicely spaced sites; it is not designed to guarantee strict non-overlap in all cases (though overlap is uncommon for moderate sizes).

## Summary: when to use which

- **Random**: general-purpose, classic random sampling, non-overlapping.
- **Tiled**: systematic grid *cells*; great for “choose random cells from a tiling”.
- **Systematic**: classic lattice-based systematic sampling.
- **Transect**: teaching/field designs with transects.
- **Voronoi**: irregular domains where you want good spatial coverage and spacing.

## Reproducibility tip

All schemes are stochastic to some degree (even the systematic scheme can depend on inferred grid dimensions); for reproducible figures, set a seed immediately before each placement call.
