---
title: "spesim recipes: environmental gradients"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{spesim recipes: environmental gradients}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

This short recipe shows how to turn **environmental filtering** on and off, and
how to change **optima/tolerances** for gradient-responsive species.

## Minimal run without special gradients

```{r}
library(spesim)

P <- load_config(system.file("examples/spesim_init_basic.txt", package = "spesim"))
P$N_SPECIES <- 10
P$N_INDIVIDUALS <- 1200
P$ADVANCED_ANALYSIS <- FALSE

set.seed(P$SEED)
res0 <- run_spatial_simulation(P = P, write_outputs = FALSE)
plot_spatial_sampling(res0$domain, res0$species_dist, res0$quadrats, res0$P)
```

## Make a few species respond strongly to a gradient

Here we keep everything else the same but make the temperature-responsive species
prefer warmer areas by shifting their **optimum** upward and tightening their
**tolerance**.

```{r}
P2 <- P

# A and B respond to temperature; C to elevation; D to rainfall (from the example init)
# Push temperature optimum to the warm end and make the response sharper
P2$GRADIENT_OPTIMA <- c(temperature = 0.80, elevation = 0.50, rainfall = 0.50)
P2$GRADIENT_TOLERANCE <- c(temperature = 0.07, elevation = 0.12, rainfall = 0.12)

res1 <- run_spatial_simulation(P = P2, write_outputs = FALSE)

p1 <- plot_spatial_sampling(res1$domain, res1$species_dist, res1$quadrats, res1$P)

# Show the underlying temperature field
p2 <- plot_spatial_sampling(
  res1$domain, res1$species_dist, res1$quadrats, res1$P,
  show_gradient = TRUE,
  env_gradients = res1$env_gradients,
  gradient_type = "temperature_C"
)

p1
p2
```

## Consequences in the advanced analysis panel

Environmental filtering typically increases **spatial turnover** (beta diversity)
when species have different optima along a gradient. In the advanced panel this
often shows up as:

- a stronger **distance–decay** signal (dissimilarity increases with distance),
- a steeper **species–area** curve (more quadrats needed to capture gamma diversity),
- and changes in **occupancy–abundance** patterns (specialists can be abundant
  but occupy fewer sites).

This is consistent with classic niche-based theory: species are sorted along
environmental axes (e.g. temperature), so communities separated along the axis
share fewer species.

```{r}
# Compare advanced panels: baseline vs stronger filtering
panel0 <- generate_advanced_panel(res0) + patchwork::plot_annotation(title = "Baseline: weak/default filtering")
panel1 <- generate_advanced_panel(res1) + patchwork::plot_annotation(title = "Stronger temperature filtering")
```

```{r fig.width=8, fig.height=9}
panel0
```

```{r fig.width=8, fig.height=9}
panel1
```

## Tip: reproducibility

If you want exact repeatability for gradient-driven patterns, set the seed (via
init file or `P$SEED`) and call `set.seed()` before running.
