---
title: "spesim recipes: point processes (Poisson vs Thomas vs Strauss/Geyer)"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{spesim recipes: point processes (Poisson vs Thomas vs Strauss/Geyer)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

This recipe shows how to switch the spatial point-process model for the dominant
species (A) and for the remaining species.

## Base configuration

```{r}
library(spesim)

P <- load_config(system.file("examples/spesim_init_complete.txt", package = "spesim"))
P$N_SPECIES <- 10
P$N_INDIVIDUALS <- 1200
P$ADVANCED_ANALYSIS <- FALSE

# Keep interactions off for clarity
P$INTERACTION_RADIUS <- 0
P$INTERACTIONS_EDGELIST <- NULL
```

## Poisson: complete spatial randomness

```{r}
P1 <- P
P1$SPATIAL_PROCESS_A <- "poisson"
P1$SPATIAL_PROCESS_OTHERS <- "poisson"

set.seed(P$SEED)
res_pois <- run_spatial_simulation(P = P1, write_outputs = FALSE, interactions_print = FALSE)
plot_spatial_sampling(res_pois$domain, res_pois$species_dist, res_pois$quadrats, res_pois$P)
```

## Thomas clustering for the dominant species

```{r}
P2 <- P
P2$SPATIAL_PROCESS_A <- "thomas"
P2$A_PARENT_INTENSITY <- NA
P2$A_MEAN_OFFSPRING <- 10
P2$A_CLUSTER_SCALE <- 1.0
P2$SPATIAL_PROCESS_OTHERS <- "poisson"

res_th <- run_spatial_simulation(P = P2, write_outputs = FALSE, interactions_print = FALSE)
plot_spatial_sampling(res_th$domain, res_th$species_dist, res_th$quadrats, res_th$P)
```

## Mild inhibition for non-dominants (Strauss)

```{r}
P3 <- P2
P3$SPATIAL_PROCESS_OTHERS <- "strauss"
P3$OTHERS_R <- 1.0
P3$OTHERS_S <- 0.5

res_st <- run_spatial_simulation(P = P3, write_outputs = FALSE, interactions_print = FALSE)
plot_spatial_sampling(res_st$domain, res_st$species_dist, res_st$quadrats, res_st$P)
```

## Consequences in the advanced analysis panel

Different point processes correspond to different *mechanisms* of spatial
structure:

- **Poisson** is a null model (complete spatial randomness).
- **Thomas clustering** is a proxy for limited dispersal or patchy habitat
  suitability, producing aggregated individuals.
- **Strauss inhibition** approximates territoriality/space limitation, producing
  repulsion among points.

In the advanced panel, these differences often show up as:

- **distance–decay**: clustering can increase dissimilarity among quadrats,
  strengthening decay,
- **species–area**: more spatial structure can increase observed beta diversity,
  steepening the curve,
- **rarefaction**: strong aggregation can reduce per-quadrat richness at small
  sample sizes (more uneven encounter rates).

Below we compare the advanced panel between Poisson, Thomas (A clustered), and
Thomas + Strauss (others inhibited).

```{r}
p_pois <- generate_advanced_panel(res_pois) + patchwork::plot_annotation(title = "Poisson (CSR)")
p_th   <- generate_advanced_panel(res_th)   + patchwork::plot_annotation(title = "Thomas clustering for A")
p_st   <- generate_advanced_panel(res_st)   + patchwork::plot_annotation(title = "Thomas (A) + Strauss inhibition (others)")
```

```{r fig.width=8, fig.height=9}
p_pois
```

```{r fig.width=8, fig.height=9}
p_th
```

```{r fig.width=8, fig.height=9}
p_st
```


## Notes

- In teaching, it often helps to hold everything else constant and vary one
  process knob at a time.
- If you want a more complete set of diagnostics, run
  `spesim_demo("advanced", make_plots = TRUE)` and inspect the advanced panel.
