---
title: "spesim workflow: basic → intermediate → advanced"
output:
  rmarkdown::html_vignette:
    toc: true
    highlight: null
vignette: >
  %\VignetteIndexEntry{spesim workflow: basic → intermediate → advanced}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

This vignette gives a structured overview of **what spesim can do**, arranged as a
progression from **basic** (fast, minimal outputs) through **intermediate**
(common teaching / exploration workflows) to **advanced** (full reporting and
diagnostic panels).

It is meant as a *map* of the package: which functions are involved, what inputs
and outputs to expect, and which “knobs” to turn at each stage.

## Big picture workflow

Below is a compact workflow diagram showing the main pipeline.

```text
┌──────────────────────────────┐
│ 1) Configuration             │
│   - load_config(init_file)   │
│   - or tweak list P in R     │
└───────────────┬──────────────┘
                │
                v
┌──────────────────────────────┐
│ 2) Run simulation            │
│   spesim_run()               │
│   - creates domain (or use)  │
│   - simulates individuals    │
│   - places quadrats          │
│   - builds matrices          │
└───────────────┬──────────────┘
                │
                v
┌──────────────────────────────────────────────────────┐
│ 3) Primary outputs (returned in-memory; optional disk)│
│   res$species_dist   (individual points, with species)│
│   res$quadrats       (quadrat polygons)               │
│   res$abund_matrix   (site × species table)           │
│   res$env_gradients  (gridded env fields)             │
│   res$site_coords    (quadrat centroids)              │
└───────────────┬──────────────────────────────────────┘
                │
                ├───────────── Basic: maps + matrices
                │
                ├───────────── Intermediate: derived analyses
                │                 - rank–abundance
                │                 - occupancy–abundance
                │                 - SAR
                │                 - distance–decay
                │                 - rarefaction
                │
                └───────────── Advanced: panels + report
                                  - generate_advanced_panel(res)
                                  - generate_full_report(res)
```

## Three levels of use

### Level 1 — Basic: fast, minimal, robust

**Goal:** run a small simulation quickly and get the essential objects back for
plotting and teaching.

**What you typically do**

```{r}
library(spesim)

# Start from a known-working example init file
init <- system.file("examples/spesim_init_basic.txt", package = "spesim")
P <- load_config(init)

# Keep it fast (in-memory only)
P$ADVANCED_ANALYSIS <- FALSE
res <- spesim_run(P, write_outputs = FALSE, seed = P$SEED)

names(res)
```

**Useful objects at this level**

- `res$domain` — sampling domain polygon
- `res$species_dist` — individuals as sf POINTS with a `species` column
- `res$quadrats` — quadrat polygons with `quadrat_id`
- `res$abund_matrix` — site × species abundance table

**Basic visualisations**

```{r}
# Domain + quadrats
plot_quadrats(res$domain, res$quadrats, title = "Quadrats")
```

### Level 2 — Intermediate: typical analyses + teaching plots

**Goal:** use the returned objects to compute common community-ecology summaries.

These are the same components that appear in the “advanced panel”, but here you
can compute/plot them separately or integrate them into lessons.

Typical intermediate steps:

- **Rank–abundance** (SAD):
  - `calculate_rank_abundance(res$species_dist, res$P)`
  - `plot_rank_abundance(...)`
- **Occupancy–abundance**:
  - `calculate_occupancy_abundance(res$abund_matrix)`
  - `plot_occupancy_abundance(...)`
- **Species–area curve (SAR)**:
  - `calculate_species_area(res$abund_matrix)`
  - `plot_species_area(...)`
- **Distance–decay**:
  - `calculate_distance_decay(res$abund_matrix, res$site_coords)`
  - `plot_distance_decay(...)`
- **Rarefaction**:
  - `calculate_rarefaction(res$abund_matrix)`
  - `plot_rarefaction(...)`

Example (one intermediate product):

```{r}
ra <- calculate_rank_abundance(res$species_dist, res$P)
plot_rank_abundance(ra)
```

### Level 3 — Advanced: full diagnostics + report

**Goal:** generate a compact **multi-plot diagnostic panel** and a
**human-readable report**.

There are two ways to do this:

1. **In-memory** (useful for notebooks and teaching):

```{r, eval=FALSE}
panel <- generate_advanced_panel(res)
print(panel)
cat(generate_full_report(res))
```

2. **Write to disk** (useful for “run it and inspect outputs” workflows):

```{r, eval=FALSE}
# This will write timestamped outputs under out/demo_YYYYMMDD... etc.
res2 <- spesim_run(P, write_outputs = TRUE, output_prefix = "out/demo", seed = P$SEED)
```

## What features live where?

### Configuration options

- “Init file” workflow: `load_config("path/to/init.txt")`
- In-memory workflow: `P <- load_config(example); P$N_SPECIES <- 10; ...`

Key configuration families:

- **Community size + SAD:** `N_SPECIES`, `N_INDIVIDUALS`, `DOMINANT_FRACTION`, `FISHER_ALPHA`, `FISHER_X`
- **Environmental filtering:** `GRADIENT_*` plus `SAMPLING_RESOLUTION`, `ENVIRONMENTAL_NOISE`
- **Point processes:** `SPATIAL_PROCESS_A`, `SPATIAL_PROCESS_OTHERS` and their parameters
- **Interactions:** `INTERACTION_RADIUS`, `INTERACTIONS_EDGELIST` or `INTERACTIONS_FILE`
- **Sampling design:** `SAMPLING_SCHEME` and scheme-specific parameters

### Sampling design: quadrat placement

- random: `place_quadrats()`
- tiled: `place_quadrats_tiled()`
- systematic: `place_quadrats_systematic()`
- transect: `place_quadrats_transect()`
- Voronoi-based: `place_quadrats_voronoi(show_voronoi=TRUE)` for teaching/visualisation

See the dedicated vignette: **“Quadrat placement schemes in spesim”**.

## Teaching progression

A suggested progression for a lab/class:

1. **Basic run** (small, fast) → inspect `res$species_dist`, `res$quadrats`, `res$abund_matrix`.
2. Turn on **gradients** → show how optima/tolerance shapes spatial pattern.
3. Compare **sampling schemes** → show how design changes the observed matrix.
4. Turn on **interactions** → show how neighbourhood effects alter co-occurrence.
5. Run the **advanced panel** → interpret each diagnostic.
6. Read the **full report** → emphasise reproducibility and “what happened?” narratives.

## Reproducibility notes

- `load_config()` sets a seed (from the init file) for reproducibility.
- Some intermediate analyses (e.g., SAR permutations) use randomness internally;
  set `set.seed(...)` before calling if you want exact repeatability.
