---
title: "spesim recipes: interspecific interactions"
output:
  rmarkdown::html_vignette:
    toc: true
    highlight: null
vignette: >
  %\VignetteIndexEntry{spesim recipes: interspecific interactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

This recipe shows how to specify simple **neighbour effects** using
`INTERACTIONS_EDGELIST`.

Values:

- `< 1` suppressive (competition)
- `> 1` facilitative
- `= 1` neutral

## Baseline without interactions

```{r}
library(spesim)

P <- load_config(system.file("examples/spesim_init_basic.txt", package = "spesim"))
P$N_SPECIES <- 10
P$N_INDIVIDUALS <- 2000
P$ADVANCED_ANALYSIS <- FALSE

# Turn interactions off
P$INTERACTION_RADIUS <- 0
P$INTERACTIONS_EDGELIST <- NULL

set.seed(P$SEED)
res0 <- run_spatial_simulation(P = P, write_outputs = FALSE, interactions_print = FALSE)
plot_spatial_sampling(res0$domain, res0$species_dist, res0$quadrats, res0$P)
```

## Add a few directed interaction rules

Here, species A suppresses B..D locally, and species C facilitates A.

```{r}
P2 <- P
P2$INTERACTION_RADIUS <- 2
P2$INTERACTIONS_EDGELIST <- c(
  "A,B-D,0.7",
  "C,A,1.3"
)

res1 <- run_spatial_simulation(P = P2, write_outputs = FALSE, interactions_print = TRUE)

# Visualise the outcome
plot_spatial_sampling(res1$domain, res1$species_dist, res1$quadrats, res1$P)
```

## Consequences in the advanced analysis panel

Neighbour effects can change **co-occurrence patterns** and local abundance,
which in turn changes what you observe in quadrats.

In ecological terms, competition (coefficients < 1) can create **spatial
segregation** and reduce local richness in neighbourhoods dominated by a strong
competitor, while facilitation (> 1) can increase local co-occurrence.

In the advanced panel, interaction effects can manifest as:

- shifts in **occupancy–abundance** (suppressed species become rarer and occupy
  fewer sites),
- changes in **distance–decay** (if interactions amplify patchiness),
- and altered **rarefaction** curves (effective evenness/richness per site).

```{r}
panel0 <- generate_advanced_panel(res0) + patchwork::plot_annotation(title = "Baseline: no interactions")
panel1 <- generate_advanced_panel(res1) + patchwork::plot_annotation(title = "With neighbour interactions")
```

```{r fig.width=8, fig.height=9}
panel0
```

```{r fig.width=8, fig.height=9}
panel1
```

## Notes

- Rules are **directed** (`A -> B` need not equal `B -> A`).
- The global `INTERACTION_RADIUS` controls the neighbourhood size.
- Keep `INTERACTION_RADIUS = 0` to disable the modifier.
