---
title: "spesim recipes: interspecific interactions"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{spesim recipes: interspecific interactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

This recipe shows how to specify simple **neighbour effects** using
`INTERACTIONS_EDGELIST`.

Values:

- `< 1` suppressive (competition)
- `> 1` facilitative
- `= 1` neutral

## Baseline (no interactions)

```{r}
library(spesim)

P <- load_config(system.file("examples/spesim_init_basic.txt", package = "spesim"))
P$N_SPECIES <- 10
P$N_INDIVIDUALS <- 2000
P$ADVANCED_ANALYSIS <- FALSE

# Turn interactions off
P$INTERACTION_RADIUS <- 0
P$INTERACTIONS_EDGELIST <- NULL

res0 <- run_spatial_simulation(P = P, write_outputs = FALSE, interactions_print = FALSE)
plot_spatial_sampling(res0$domain, res0$species_dist, res0$quadrats, res0$P)
```

## Add a few directed interaction rules

Here, species A suppresses B..D locally, and species C facilitates A.

```{r}
P2 <- P
P2$INTERACTION_RADIUS <- 2
P2$INTERACTIONS_EDGELIST <- c(
  "A,B-D,0.7",
  "C,A,1.3"
)

res1 <- run_spatial_simulation(P = P2, write_outputs = FALSE, interactions_print = TRUE)

# Visualise the outcome
plot_spatial_sampling(res1$domain, res1$species_dist, res1$quadrats, res1$P)
```

## Notes

- Rules are **directed** (`A -> B` need not equal `B -> A`).
- The global `INTERACTION_RADIUS` controls the neighbourhood size.
- Keep `INTERACTION_RADIUS = 0` to disable the modifier.
