---
title: "Method testing with spesim"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Method testing with spesim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)

library(spesim)
library(dplyr)
library(ggplot2)
```

spesim is designed for **teaching** and **method testing**.

Method testing means: you keep the *underlying truth* fixed (the simulated community + spatial process), and vary the **sampling design** or **analysis choices**, then check how your conclusions change.

This vignette shows a simple, reproducible workflow:

1) Run a baseline simulation.
2) Re-run the **same truth** under different sampling schemes.
3) Compare (i) what you would conclude from the usual plots/tables and (ii) what the **audit layer** says you actually got.

## 1. Start from a baseline parameter set

We start from the packaged init file and then modify only a few parameters.

```{r}
`%||%` <- function(a, b) if (!is.null(a)) a else b

init_file <- system.file("examples/spesim_init_basic.txt", package = "spesim")
if (identical(init_file, "")) {
  # Useful when rendering from the source tree without a prior install.
  # During local rendering, the working directory is typically `vignettes/`.
  cand <- c(
    file.path("..", "inst", "examples", "spesim_init_basic.txt"),
    file.path("..", "spesim_init_basic.txt")
  )
  init_file <- cand[file.exists(cand)][1]
}

P0 <- load_config(init_file)

# Keep this vignette reasonably fast.
P0$N_INDIVIDUALS <- 1500L
P0$N_QUADRATS <- 25L

# Ensure we have filtering species (default init already does), and keep the report light.
P0$ADVANCED_ANALYSIS <- FALSE

# Fix seed so the *truth* is reproducible.
P0$SEED <- 123L
```

## 2. A helper to run the same truth under different sampling schemes

`spesim_method_test()` is a convenience wrapper that returns:
- `res`: the simulation result
- `audit`: a lightweight audit of the realised regime
- optional standard plots and tables

```{r}
run_scheme <- function(P, scheme) {
  P2 <- P
  P2$SAMPLING_SCHEME <- scheme

  # Keep audit deterministic and dependency-minimal for vignettes.
  spesim_method_test(P = P2, diagnostics = "nn", make_tables = TRUE, make_plots = FALSE)
}

schemes <- c("random", "systematic", "tiled", "transect", "voronoi")
outs <- setNames(lapply(schemes, function(s) run_scheme(P0, s)), schemes)
```

## 3. Compare the sampling audits (what did the sampling design *actually* do?)

Sampling schemes differ in how much they are constrained by the boundary and by the non-overlap rule.

```{r}
sampling_summary <- bind_rows(lapply(names(outs), function(s) {
  a <- outs[[s]]$audit
  data.frame(
    scheme = s,
    n_requested = a$sampling$n_requested %||% NA_integer_,
    n_returned  = a$sampling$n_returned  %||% NA_integer_,
    reject_boundary = a$sampling$reject_boundary %||% NA_integer_,
    reject_overlap  = a$sampling$reject_overlap  %||% NA_integer_,
    safe_area_fraction = a$sampling$safe_area_fraction %||% NA_real_
  )
}))

sampling_summary
```

Interpretation tips:

- `safe_area_fraction` is a conservative estimate of how much of the domain is a feasible sampling frame for quadrat *centres*.
- High `reject_boundary` or low `safe_area_fraction` means your design is effectively sampling the interior only.

## 4. Compare how the same truth looks under different schemes

A small, practical comparison is to look at a few common summaries.

### 4.1 Mean richness and abundance per quadrat

```{r}
quadrat_summaries <- bind_rows(lapply(names(outs), function(s) {
  res <- outs[[s]]$res
  abund <- res$abund_matrix[, setdiff(colnames(res$abund_matrix), "site"), drop = FALSE]

  data.frame(
    scheme = s,
    mean_richness = mean(rowSums(abund > 0)),
    mean_abundance = mean(rowSums(abund))
  )
}))

quadrat_summaries
```

### 4.2 Distance-decay (beta diversity vs distance)

We compute a distance-decay table for each scheme and compare the *fitted slope*.

```{r}
slopes <- bind_rows(lapply(names(outs), function(s) {
  res <- outs[[s]]$res
  dd <- calculate_distance_decay(res$abund_matrix, res$site_coords)

  # Simple linear trend (teaching-level summary): dissimilarity ~ distance
  if (nrow(dd) == 0) {
    return(data.frame(scheme = s, slope = NA_real_, r2 = NA_real_))
  }
  fit <- lm(Dissimilarity ~ Distance, data = dd)

  data.frame(
    scheme = s,
    slope = unname(coef(fit)["Distance"]),
    r2 = summary(fit)$r.squared
  )
}))

slopes
```

**Method-testing point:** If different sampling schemes change the slope materially, your inference about spatial turnover may be design-sensitive.

## 5. Use the audit layer to prevent over-interpretation

The audit tells you whether the regime you *think* you requested is visible in the sampled data.

For example, environmental filtering checks can be noisy when a species is sparse.

```{r}
filtering_summary <- bind_rows(lapply(names(outs), function(s) {
  f <- outs[[s]]$audit$filtering
  if (nrow(f) == 0) return(data.frame())

  # Keep a compact view for the vignette
  f %>%
    mutate(scheme = s) %>%
    select(scheme, species, gradient, n_sites, occupancy_sites, rho, slope, qualitative)
}))

filtering_summary
```

Interpretation tips:

- Prefer filtering interpretations when `occupancy_sites` is not tiny.
- Treat `too_sparse_to_assess` as a guardrail: it means *your sampled data don’t support a strong statement*.

## 6. What to do next

To turn this into a real method-testing exercise:

- Repeat the comparison under different domain shapes (e.g., very concave polygons).
- Vary quadrat size and number of quadrats.
- Hold sampling fixed and vary the spatial process (Poisson vs Strauss vs Thomas).
- Use `spesim_audit(..., diagnostics = c("nn", "spatstat"))` when `spatstat` is installed for edge-corrected point-process diagnostics.
