---
title: "Validation and sanity checks"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\VignetteIndexEntry{Validation and sanity checks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

This article is a **confidence-building checklist**. It shows simple checks that
help you confirm that a spesim run is producing the patterns you think you asked
for.

## 1) Species–abundance distribution (SAD)

The rank–abundance plot compares realised counts to the target Fisher log-series
settings.

```{r, eval=FALSE}
library(spesim)

P <- load_config(system.file("examples/spesim_init_basic.txt", package = "spesim"))
res <- spesim_run(P, write_outputs = FALSE, seed = 77)

ra <- calculate_rank_abundance(res$species_dist)
plot_rank_abundance(ra)
```

What to look for:

- Realised counts are close to the theoretical curve (on log-scale) when sample
  sizes are large.
- Big departures usually indicate strong filtering / strong interactions / very
  small sample sizes.

## 2) Quadrat placement geometry

Always sanity-check the design visually.

```{r, eval=FALSE}
plot_quadrats(res$domain, res$quadrats, title = "Quadrat layout")
```

## 3) Distance–decay behaves sensibly

```{r, eval=FALSE}
dd <- calculate_distance_decay(res$abund_matrix, res$site_coords)
plot_distance_decay(dd)
```

What to look for:

- With strong gradients or dispersal limitation-like structure, dissimilarity
  should generally increase with distance.
- Under CSR + no filtering, distance–decay should be weak.

## 4) Optional: point-process diagnostics with spatstat

spesim aims to stay lightweight, but if you have **spatstat** installed you can
use it to validate clustering/inhibition patterns (e.g., Ripley’s K, pair
correlation \(g(r)\)).

### A) The built-in audit (recommended)

If `spatstat.geom` and `spatstat.explore` are available, `spesim_audit()` can
compute **edge-corrected** summaries inside the true irregular window:

```{r, eval=FALSE}
a <- spesim_audit(res, diagnostics = c("nn", "spatstat"))
print(a)
```

### B) Manual workflow (power users)

```{r, eval=FALSE}
if (requireNamespace("spatstat.geom", quietly = TRUE) &&
    requireNamespace("spatstat.explore", quietly = TRUE)) {

  library(spatstat.geom)
  library(spatstat.explore)

  # Example: diagnose the dominant species pattern
  dom <- res$species_dist[res$species_dist$species == "A", ]

  # Convert to spatstat ppp in the true irregular window
  win <- as.owin(sf::st_as_sf(sf::st_union(res$domain)))
  xy <- sf::st_coordinates(dom)
  X <- ppp(xy[,1], xy[,2], window = win)

  plot(Kest(X, correction = "border"), main = "Ripley's K (border): dominant species")
  plot(pcf(X, correction = "Ripley"), main = "Pair correlation g(r): dominant species")
}
```

## 5) Reproducibility note (vegan)

Some summaries (e.g., `vegan::specaccum(method = "random")`) use internal random
permutations.

- If you want deterministic repeats, set a seed immediately beforehand:

```{r, eval=FALSE}
set.seed(res$P$SEED)
# then regenerate SAR/rarefaction plots
```
