---
title: "Performance notes"
output:
  rmarkdown::html_vignette:
    toc: true
    highlight: null
vignette: >
  %\VignetteIndexEntry{Performance notes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

spesim is intended for **teaching** and **method testing**, where you usually want runs to be *interactive* (seconds, not minutes).

This vignette gives conservative rules of thumb for keeping runtime and memory use reasonable.

## What tends to dominate runtime

1) **Neutral-model simulation loop** (when `MODEL_FAMILY = "hybrid"` or `"neutral"`).
   - The birth–death–dispersal loop runs for `NEUTRAL_MAX_STEPS` iterations.
   The inner loop is written in R; cost scales with `N_INDIVIDUALS` and the
   number of steps to convergence. As of v0.5.2 the main per-step costs are
   the C++ point-in-polygon check (`pip_cpp`) and the R interpreter overhead
   of the loop body itself — the environmental acceptance and displacement
   sampling steps are now fully vectorised over the candidate batch.

2) **Spatial intersections** (`sf::st_intersects`) between individuals and quadrats.
   - Roughly scales with the number of individuals × number of quadrats.

3) **Environmental grid resolution** (the `env_gradients` raster-like table).
   - Higher resolution increases memory and slows operations that summarise environment per quadrat.

4) **Optional analyses** (reports, panels, diversity calculations).
   - `ADVANCED_ANALYSIS = TRUE` is intentionally heavier.

## Safe interactive ranges (rules of thumb)

These are not hard limits—just settings that tend to work well on a laptop.

### Individuals

- **Good default:** `N_INDIVIDUALS` ≈ 1,000–5,000
- **Still workable for method testing:** up to ~20,000 (expect slower plotting/intersections)

### Quadrat sampling

- **Good default:** `N_QUADRATS` ≈ 10–50
- Very large `N_QUADRATS` increases intersection work and can make some analyses noisy.

### Environmental grid

- `SAMPLING_RESOLUTION` is the number of grid cells per side for the environmental surface.

- **Good default:** 30–80
- Above ~150 you should expect larger memory use and slower quadrat-environment summaries.

### Advanced outputs

- For interactive work, start with:

```{r, eval=FALSE}
P$ADVANCED_ANALYSIS <- FALSE
x <- spesim_method_test(P = P, make_plots = TRUE)
```

- Enable advanced reporting once your parameters are stable:

```{r, eval=FALSE}
P$ADVANCED_ANALYSIS <- TRUE
res <- spesim_run(P, write_outputs = TRUE, seed = P$SEED)
cat(generate_full_report(res, include_audit = TRUE), sep = "\n")
```

## Practical tips

- When comparing sampling schemes, keep the *truth* fixed (same seed) and change only sampling parameters.
- Prefer running many small replicates over one enormous run.
- If a sampling scheme returns **very few quadrats**, some downstream summaries may be undefined or uninformative (distance–decay requires ≥2 sites, etc.).

## If you need to go bigger

- Reduce plotting.
- Increase `N_INDIVIDUALS` gradually.
- Consider using the fast point-process engines when available.

## Engine performance history

Internal profiling (N = 1,500, S = 15, hybrid model) shows the following
cumulative improvements since v0.5.1:

| Release | Change | Approx. wall time |
|---------|--------|-------------------|
| v0.5.1 | C++ PIP engine + vectorised domain checks | 17,690 ms |
| v0.5.2 | Vectorised env-acceptance + displacement sampling | **2,600 ms** |

The residual cost is now dominated by the C++ PIP engine and the R interpreter
overhead of the simulation loop itself, both of which are at or near their
practical floor without a full Rcpp port of the inner loop.
