---
title: "spesim recipes: point processes (and the neutral-theory SAD helper)"
output:
  rmarkdown::html_vignette:
    toc: true
    highlight: null
vignette: >
  %\VignetteIndexEntry{spesim recipes: point processes (and the neutral-theory SAD helper)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

This recipe shows how to:

1. switch the **spatial point-process** model for the dominant species (A) and for the remaining species,
2. use the **neutral-theory SAD helper** (`SAD_MODEL = "zsm"`) *as currently implemented*.

The key idea is that these are **orthogonal knobs**:

- `SAD_MODEL` controls *how many individuals each species gets* (the rank–abundance curve),
- `SPATIAL_PROCESS_*` controls *where individuals are placed in space*.

## Base configuration

```{r}
library(spesim)
library(ggplot2)

P0 <- load_config(system.file("examples/spesim_init_complete.txt", package = "spesim"))

# Keep the vignette fast but still visually clear.
P0$N_SPECIES <- 22
P0$N_INDIVIDUALS <- 2400

# Make distance–decay curves more interpretable by using a regular quadrat layout.
P0$SAMPLING_SCHEME <- "tiled"
P0$N_QUADRATS <- 36
P0$QUADRAT_SIZE_OPTION <- "medium"

# Turn off environmental filtering to keep this vignette focused on
# point-process and SAD effects (not gradient-driven turnover).
P0$GRADIENT_SPECIES <- character(0)
P0$GRADIENT_ASSIGNMENTS <- character(0)
P0$GRADIENT_OPTIMA <- NULL
P0$GRADIENT_TOLERANCE <- NULL

# Keep interactions off for clarity
P0$INTERACTION_RADIUS <- 0
P0$INTERACTIONS_FILE <- NULL
P0$INTERACTIONS_EDGELIST <- NULL

# A tiny helper to run a case and build its advanced panel.
run_case <- function(P, title) {
  res <- spesim_run(P, write_outputs = FALSE, seed = P$SEED, interactions_print = FALSE)
  panel <- generate_advanced_panel(res) + patchwork::plot_annotation(title = title)
  list(res = res, panel = panel)
}

rank_df <- function(res, label) {
  ra <- calculate_rank_abundance(res$species_dist, res$P)
  ra <- subset(ra, Source == "Observed")
  ra$Scenario <- label
  ra
}

decay_df <- function(res, label) {
  dd <- calculate_distance_decay(res$abund_matrix, res$site_coords)
  dd$Scenario <- label
  dd
}
```

## Neutral theory in spesim (what it can and cannot do)

spesim’s **neutral theory option** is currently implemented as a **species–abundance distribution (SAD) helper**:

- Set `SAD_MODEL = "zsm"` and provide `ZSM_THETA` (and optionally `ZSM_M`).
- spesim then generates an **abundance vector** (counts per species) consistent with a neutral-theory sampling model.

### What it *can* do

- Generate **neutral-like rank–abundance curves** (SADs), controlled mainly by `ZSM_THETA`.
  - Smaller `theta` → stronger dominance / longer tail.
  - Larger `theta` → more even communities.
- Currently, spesim uses a **theta-only Ewens sampler** (Chinese restaurant / Ewens sampling formula).
- `ZSM_M` is accepted for forward compatibility, but is **currently ignored**.
- If you need a mechanistic neutral model with immigration / metacommunity coupling, use **untb** directly; spesim’s neutral option is deliberately a SAD helper.

### What it *cannot* do (yet)

- It is **not** a full mechanistic neutral simulation engine:
  - no births/deaths through time,
  - no explicit dispersal as a population process,
  - no dynamic metacommunity/local community coupling.
- It does **not** create spatial structure by itself.
  - Spatial patterns still come from `SPATIAL_PROCESS_A` / `SPATIAL_PROCESS_OTHERS` (Poisson/Thomas/Strauss/Geyer).

The practical takeaway: use `SAD_MODEL = "zsm"` for **neutral-like abundance structure**, and pair it with a point-process choice to get the **spatial patterns** you want.

## Example A: SAD differences (Fisher vs neutral-theory ZSM)

Here we hold the spatial process fixed at **Poisson** (CSR) so differences in the **SAD plots** are driven by `SAD_MODEL`.

```{r}
P_sad <- P0
P_sad$SPATIAL_PROCESS_A <- "poisson"
P_sad$SPATIAL_PROCESS_OTHERS <- "poisson"

# Fisher baseline
P_fisher <- P_sad
P_fisher$SAD_MODEL <- "fisher"
P_fisher$DOMINANT_FRACTION <- 0.30

# Neutral-like SADs via zsm
P_zlow <- P_sad
P_zlow$SAD_MODEL <- "zsm"
P_zlow$ZSM_THETA <- 1.5
P_zlow$ZSM_M <- NA_real_

P_zhigh <- P_zlow
P_zhigh$ZSM_THETA <- 50

case_fisher <- run_case(P_fisher, "Poisson + Fisher SAD")
case_zlow   <- run_case(P_zlow,   "Poisson + ZSM (theta = 1.5)")
case_zhigh  <- run_case(P_zhigh,  "Poisson + ZSM (theta = 50)")
```

### Rank–abundance (overlay)

```{r}
ra_all <- rbind(
  rank_df(case_fisher$res, "Fisher"),
  rank_df(case_zlow$res,   "ZSM θ=1.5"),
  rank_df(case_zhigh$res,  "ZSM θ=50")
)

ggplot(ra_all, aes(Rank, Abundance, colour = Scenario)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_log10() +
  labs(
    title = "Rank–abundance curves: Fisher vs neutral-theory helper",
    y = "Abundance (log10 scale)",
    x = "Rank (1 = most abundant)"
  ) +
  theme_minimal(base_size = 12)
```

### Full advanced panels

```{r fig.width=8, fig.height=9}
case_fisher$panel
```

```{r fig.width=8, fig.height=9}
case_zlow$panel
```

```{r fig.width=8, fig.height=9}
case_zhigh$panel
```

## Example B: Spatial differences (distance–decay) under a fixed neutral SAD

Now we fix the SAD model to **neutral-theory ZSM** and vary only the **point-process**.

To make the contrast obvious, we choose a strongly structured placement:

- Dominant species **A**: tight **Thomas clustering** (many offspring, small sigma)
- Other species: strong **Strauss inhibition** (repulsion)

```{r}
P_neutral <- P0
P_neutral$SAD_MODEL <- "zsm"
P_neutral$ZSM_THETA <- 10
P_neutral$ZSM_M <- NA_real_

# CSR baseline
P_neutral_pois <- P_neutral
P_neutral_pois$SPATIAL_PROCESS_A <- "poisson"
P_neutral_pois$SPATIAL_PROCESS_OTHERS <- "poisson"

# Strong spatial structure
P_neutral_struct <- P_neutral
P_neutral_struct$SPATIAL_PROCESS_A <- "thomas"
P_neutral_struct$A_PARENT_INTENSITY <- NA
P_neutral_struct$A_MEAN_OFFSPRING <- 25
P_neutral_struct$A_CLUSTER_SCALE <- 0.45

P_neutral_struct$SPATIAL_PROCESS_OTHERS <- "strauss"
P_neutral_struct$OTHERS_R <- 1.2
P_neutral_struct$OTHERS_S <- 0.15

case_neutral_pois   <- run_case(P_neutral_pois,   "ZSM (theta=10) + Poisson (CSR)")
case_neutral_struct <- run_case(P_neutral_struct, "ZSM (theta=10) + Thomas (A) + Strauss (others)")
```

### Distance–decay (overlay)

```{r}
dd_all <- rbind(
  decay_df(case_neutral_pois$res,   "Poisson"),
  decay_df(case_neutral_struct$res, "Thomas + Strauss")
)

ggplot(dd_all, aes(Distance, Dissimilarity, colour = Scenario)) +
  geom_point(alpha = 0.25) +
  geom_smooth(se = FALSE, linewidth = 1.2, method = "loess", span = 0.75) +
  labs(
    title = "Distance–decay under the same neutral SAD",
    subtitle = "Spatial structure changes turnover with distance; SAD model is held constant",
    x = "Inter-quadrat distance",
    y = "Community dissimilarity (Bray–Curtis on presence/absence)"
  ) +
  theme_minimal(base_size = 12)
```

### Full advanced panels

```{r fig.width=8, fig.height=9}
case_neutral_pois$panel
```

```{r fig.width=8, fig.height=9}
case_neutral_struct$panel
```

## Notes / tips

- If you want *neutral-like abundance* **and** *neutral-like dispersal/clustering*, you currently need to **combine**:
  - `SAD_MODEL = "zsm"` (abundances) with
  - a spatial process such as `thomas` (placement).
- If the distance–decay looks noisy, increase `N_QUADRATS` and/or reduce quadrat size.
- For pedagogy, it helps to vary **one knob at a time** (SAD vs point process), then show how they interact.
