---
title: "Even and uneven communities: a study in species diversity"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    highlight: null
vignette: >
  %\VignetteIndexEntry{Even and uneven communities: a study in species diversity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
library(spesim)
```

## Motivation

One of the clearest illustrations of what diversity metrics actually measure
comes from two contrasting communities described by David Zelený in his
[Analysis of Community Ecology Data in R](https://www.davidzeleny.net/anadat-r/doku.php/en:diversity_analysis)
course materials:

- **Community A** — 20 species, each represented by exactly the same number
  of individuals (perfectly even).
- **Community B** — 20 species, but one dominant species holds 81% of all
  individuals while the remaining 19 species each hold 1%.

Both communities have the same *species richness* (S = 20), yet they are
ecologically very different. Community A has high **evenness**: if you sample
one individual at random, every species is equally likely to appear. Community
B has near-zero evenness: almost every individual you pick will be the
dominant species.

Zelény uses these communities to motivate the need for diversity metrics
beyond richness — Shannon H', Simpson D, and the concept of
**encounterability** (the probability that the next individual you encounter
belongs to a species you have not yet seen).

Here we use **spesim** to simulate both communities in a spatial setting and
compare them through rank–abundance curves, occupancy–abundance plots,
species–area curves, distance–decay, and rarefaction.

---

## Ecological rationale

### Community A — perfectly even

In ecological theory, the **broken-stick model** (MacArthur 1957) describes a
community where S species divide a fixed resource axis into S segments by
placing S − 1 random "breaks". Each species occupies one segment, so segment
lengths — and thus species abundances — are as evenly distributed as a random
partition allows.

A perfectly even community pushes this further: all species have *exactly* the
same proportion. This can arise when:

- Niche differentiation is complete (each species occupies a distinct,
  equally-sized niche),
- Competition is symmetric (no species has a competitive advantage),
- Demographic stochasticity has had time to average out across a large
  community.

In practice, such perfect evenness is rare, but it serves as a useful null
model and a teaching benchmark.

### Community B — highly uneven

At the opposite extreme, a community can be dominated by a single species
that monopolises most of the available resource or physical space. This
pattern is common in systems shaped by:

- Competitive exclusion (one superior competitor suppresses others),
- Priority effects (the species that arrived first pre-empts resources),
- Stress or disturbance (only the most tolerant species thrives),
- Allelopathy or other asymmetric biotic interactions.

Species B through T persist as rare species — present, but contributing
negligibly to total abundance. In sampling, rare species are easily missed
unless sample sizes are very large, which is why rarefaction and occupancy–
abundance relationships look so different from Community A.

---

## Specifying communities in spesim

The key decision is the **SAD model** (`SAD_MODEL`) and its parameters. spesim
ships several built-in SAD generators (Fisher log-series, geometric series,
broken-stick, Zipf, lognormal, Poisson–lognormal, ZSM/neutral), but for exact
control we use `SAD_MODEL = custom` together with an explicit probability
vector (`SAD_VECTOR`).

Both communities share the same structural choices that isolate the SAD as the
only variable:

| Setting | Value | Rationale |
|---------|-------|-----------|
| `MODEL_FAMILY` | `manual` | Abundances from SAD, placed by point process |
| `SPATIAL_PROCESS_A` | `poisson` | Complete Spatial Randomness for all species |
| `SPATIAL_PROCESS_OTHERS` | `poisson` | As above |
| `INTERACTION_RADIUS` | `0` | No local biotic interactions |
| No `GRADIENT_SPECIES` | — | No environmental filtering |

By removing spatial structure and environmental gradients we ensure that
**any observed differences between the two communities come purely from their
species abundance distributions**, not from space.

### Init file for Community A

```
SEED = 42
OUTPUT_PREFIX = out/community_a

N_INDIVIDUALS = 2000
N_SPECIES     = 20
MODEL_FAMILY  = manual

DOMAIN_TYPE = polygon

SAD_MODEL  = custom
SAD_VECTOR = c(rep(0.05, 20))   # all 20 species at 5% each

INTERACTION_RADIUS = 0

SPATIAL_PROCESS_A      = poisson
SPATIAL_PROCESS_OTHERS = poisson

SAMPLING_SCHEME     = random
N_QUADRATS          = 20
QUADRAT_SIZE_OPTION = medium

ADVANCED_ANALYSIS = FALSE
```

`c(rep(0.05, 20))` is valid R syntax; the config parser evaluates it directly
to a 20-element numeric vector summing to 1.0. Because the vector sums to 1,
spesim interprets it as species probabilities and draws integer abundances via
`stats::rmultinom()`.

### Init file for Community B

```
SEED = 42
OUTPUT_PREFIX = out/community_b

N_INDIVIDUALS = 2000
N_SPECIES     = 20
MODEL_FAMILY  = manual

DOMAIN_TYPE = polygon

SAD_MODEL  = custom
SAD_VECTOR = c(0.81, rep(0.01, 19))  # species A at 81%; B–T at 1% each

INTERACTION_RADIUS = 0

SPATIAL_PROCESS_A      = poisson
SPATIAL_PROCESS_OTHERS = poisson

SAMPLING_SCHEME     = random
N_QUADRATS          = 20
QUADRAT_SIZE_OPTION = medium

ADVANCED_ANALYSIS = FALSE
```

`c(0.81, rep(0.01, 19))` sums to exactly 1.0. The first element maps to
species A (the "dominant" slot in spesim's internal labelling); elements 2–20
map to species B through T.

---

## Simulation

```{r load-configs}
init_A <- system.file("examples/spesim_init_community_a.txt", package = "spesim")
init_B <- system.file("examples/spesim_init_community_b.txt", package = "spesim")

P_A <- load_config(init_A)
P_B <- load_config(init_B)
```

```{r run-A}
res_A <- spesim_run(P_A, write_outputs = FALSE, seed = P_A$SEED)
```

```{r run-B}
res_B <- spesim_run(P_B, write_outputs = FALSE, seed = P_B$SEED)
```

---

## Community maps

The maps below show all 2,000 simulated individuals coloured by species.
Notice that in Community A every species colour is roughly equally visible,
whereas in Community B one colour (species A) dominates the canvas and the
other 19 appear as scattered dots.

```{r map-A, fig.height=5}
plot_quadrats(res_A$domain, res_A$quadrats, title = "Community A — even (20 species × 5%)")
```

```{r map-B, fig.height=5}
plot_quadrats(res_B$domain, res_B$quadrats, title = "Community B — uneven (1 × 81%, 19 × 1%)")
```

---

## Advanced analysis panels

`generate_advanced_panel()` assembles five diagnostic plots from a single
simulation result.

### Community A

```{r panel-A, fig.height=9, fig.width=9}
set.seed(P_A$SEED)
generate_advanced_panel(res_A)
```

### Community B

```{r panel-B, fig.height=9, fig.width=9}
set.seed(P_B$SEED)
generate_advanced_panel(res_B)
```

---

## Interpreting the panels

### 1) Rank–abundance (SAD)

**Community A** — the curve is nearly flat. All 20 species have similar
counts; the rank ordering is an artefact of the random multinomial draw rather
than a systematic dominance gradient.

**Community B** — a steep cliff followed by a plateau. Species A sits far
above the rest; species B–T cluster near the bottom, essentially
indistinguishable from one another on the log scale.

### 2) Occupancy–abundance

Occupancy (number of quadrats in which a species is present) is plotted
against total abundance.

**Community A** — points form a loose positive cloud near the top-right: all
species are both abundant and widely distributed. High encounterability means
that even a small quadrat is likely to contain most species.

**Community B** — a bimodal scatter. Species A occupies essentially every
quadrat (high occupancy, high abundance), while species B–T cluster at the
bottom-left (low abundance, low and variable occupancy). Many rare species
will be missed by any individual quadrat.

### 3) Species–area (specaccum)

**Community A** — the accumulation curve rises steeply from the first
quadrat. Because all species are common, most species are detected in the
first few sampling units and the curve quickly levels off near S = 20.

**Community B** — the curve rises initially (picking up species A quickly),
then climbs more slowly as additional quadrats begin to capture the rare
species. The curve may not level off fully even at 20 quadrats: very rare
species (1% each, 20 individuals out of 2,000) may be absent from some
quadrats by chance.

### 4) Distance–decay

Both communities use Poisson (CSR) spatial processes, so there is no spatial
structure: community composition does not change systematically with distance.
Expect a flat, low-dissimilarity scatter in Community A (most quadrats share
most species) and a high-but-flat scatter in Community B (most quadrats share
only species A and miss the rarest species, producing high Sørensen
dissimilarity that does not increase with distance).

### 5) Rarefaction

The rarefaction curves reveal how quickly species accumulate as a function of
the number of *individuals* sampled.

**Community A** — all species-level curves rise sharply. Because every
species is at 5% of abundance, you encounter a new species frequently. The
expected richness at even a small subsample size is high.

**Community B** — the per-quadrat curves reflect sampling dominated by species
A. Most of the curve represents adding more individuals of the same species.
The curve flattens early for the dominant and only picks up rare species at
very high individual counts. The expected richness at a standard rarefaction
depth (e.g. 50 individuals) will be much lower than for Community A.

---

## Quantitative diversity summary

We can compute a brief diversity summary directly from the abundance matrix.

```{r diversity-summary}
# Drop the 'site' label column; keep only numeric species columns
num_cols_A <- sapply(res_A$abund_matrix, is.numeric)
num_cols_B <- sapply(res_B$abund_matrix, is.numeric)

abund_A <- as.matrix(res_A$abund_matrix[, num_cols_A])
abund_B <- as.matrix(res_B$abund_matrix[, num_cols_B])

shannon_A <- vegan::diversity(abund_A, index = "shannon")
shannon_B <- vegan::diversity(abund_B, index = "shannon")

pielou_J <- function(H, S) H / log(S)

S_A <- vegan::specnumber(abund_A)
S_B <- vegan::specnumber(abund_B)

cat(sprintf(
  "Community A — mean Shannon H': %.3f | mean Pielou J: %.3f | mean richness: %.1f\n",
  mean(shannon_A), mean(pielou_J(shannon_A, S_A)), mean(S_A)
))
cat(sprintf(
  "Community B — mean Shannon H': %.3f | mean Pielou J: %.3f | mean richness: %.1f\n",
  mean(shannon_B), mean(pielou_J(shannon_B, S_B)), mean(S_B)
))
```

Despite identical species richness in the pool (S = 20), the two communities
differ dramatically in Shannon H' and Pielou's J — exactly as Zelény's
examples predict. Even the *observed* richness per quadrat differs, because
the rare species in Community B are frequently absent from individual sampling
units.

---

## Key take-aways

1. **Richness alone is insufficient.** Both communities have 20 species
   in the pool, but their ecological properties are completely different.

2. **Evenness structures the entire detection process.** In Community A,
   a handful of individuals is enough to encounter most species; in Community
   B you need many more individuals to move beyond the dominant.

3. **The SAD model is the primary lever.** By fixing all spatial and
   environmental settings and changing only `SAD_VECTOR`, spesim isolates
   the effect of the abundance distribution on every downstream metric.

4. **Distance–decay is spatial, not SAD-driven.** With Poisson processes,
   both communities show no systematic distance–decay — any dissimilarity
   between quadrats is random and does not increase with distance. Introducing
   spatial clustering (Thomas process) or environmental filtering would add a
   distance–decay signal on top of the SAD effect.

---

## Further reading

- Zelený D. (n.d.). *Analysis of Community Ecology Data in R*.
  <https://www.davidzeleny.net/anadat-r/doku.php/en:diversity_analysis>
- MacArthur R.H. (1957). On the relative abundance of bird species.
  *Proceedings of the National Academy of Sciences*, **43**(3), 293–295.
- Magurran A.E. (2004). *Measuring Biological Diversity*. Blackwell.
- Chao A. & Chiu C.-H. (2016). Nonparametric estimation and comparison of
  species richness. In: *eLS*. Wiley.
