---
title: "Start here"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\VignetteIndexEntry{Start here}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

This is the quickest path through the package, depending on what you’re trying
to do.

## 1) First: understand what spesim is

- **Model card:** `vignette("spesim-model-card")`

If you’re teaching with spesim, start there: it sets expectations and prevents
over-interpretation.

## 2) Run your first simulation

- **Workflow overview:** `vignette("spesim-workflow")`

Minimal, in-memory run:

```{r, eval=FALSE}
library(spesim)

P <- load_config(system.file("examples/spesim_init_basic.txt", package = "spesim"))
res <- spesim_run(P, write_outputs = FALSE, seed = 77)

plot_spatial_sampling(res$domain, res$species_dist, res$quadrats, res$P)
```

## 3) Choose a sampling design

- **Quadrat placement schemes:** `vignette("spesim-quadrat-placement")`

## 4) Add structure (gradients, interactions, point processes)

- **Environmental gradients:** `vignette("spesim-env-gradients")`
- **Interactions:** `vignette("spesim-interactions")`
- **Point processes:** `vignette("spesim-point-processes")`

## 5) Validate that the patterns match your intent

- **Validation & sanity checks:** `vignette("spesim-validation")`

## 6) Stability / teaching materials

If you rely on specific functions in teaching content:

- **Public API (what’s stable):** `vignette("spesim-public-api")`
