---
title: "spesim: real-world constrained landscapes (Doubs + seaweed)"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{spesim: real-world constrained landscapes (Doubs + seaweed)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

This vignette shows how to use real-world coordinates and covariates from
`inst/extdata`:

- Doubs river data: `DoubsSpa.csv`, `DoubsEnv.csv`, `DoubsSpe.csv`
- Seaweed coastline data: `SeaweedSpa.csv`, `SeaweedEnv.csv`, `SeaweedSpe.csv`

It also uses pre-merged covariate files shipped in the package:

- `DoubsCovariates.csv`
- `SeaweedCovariates.csv`

## Limitation note

This is a linearized constrained-landscape MVP, not yet a full graph-theoretic
river-network engine (no explicit edge/node topology or true network
shortest-path over branches yet).

## Helpers

```{r}
library(spesim)
library(ggplot2)
library(sf)

read_with_site <- function(path) {
  x <- read.csv(path, check.names = FALSE)
  if ("" %in% names(x)) names(x)[names(x) == ""] <- "site"
  x
}

pkg_file <- function(subpath, folder = c("extdata", "examples")) {
  folder <- match.arg(folder)
  p_installed <- system.file(folder, subpath, package = "spesim")
  if (nzchar(p_installed) && file.exists(p_installed)) return(p_installed)
  candidates <- c(
    file.path("inst", folder, subpath),
    file.path("..", "inst", folder, subpath)
  )
  for (p_local in candidates) {
    if (file.exists(p_local)) return(p_local)
  }
  stop("Could not find file: ", subpath, " in ", folder)
}

make_domain_from_xy <- function(df, buffer_frac = 0.06) {
  pts <- st_as_sf(df, coords = c("x", "y"), crs = NA)
  hull <- st_convex_hull(st_union(pts))
  bb <- st_bbox(hull)
  d <- sqrt((bb["xmax"] - bb["xmin"])^2 + (bb["ymax"] - bb["ymin"])^2)
  st_as_sf(st_buffer(hull, d * buffer_frac))
}

build_linear_pos <- function(x) {
  rng <- range(x, na.rm = TRUE)
  den <- rng[2] - rng[1]
  if (!is.finite(den) || den == 0) den <- 1
  pmax(0, pmin(1, (x - rng[1]) / den))
}
```

## 1) Tie Doubs environmental data to real coordinates

`DoubsSpa.csv` provides site coordinates (`X`, `Y`). `DoubsEnv.csv` provides
site-level environmental variables. We join by site ID.

```{r}
doubs_spa <- read_with_site(pkg_file("DoubsSpa.csv", "extdata"))
doubs_env <- read_with_site(pkg_file("DoubsEnv.csv", "extdata"))
doubs_spe <- read_with_site(pkg_file("DoubsSpe.csv", "extdata"))

doubs <- merge(doubs_spa, doubs_env, by = "site", sort = TRUE)
names(doubs)[names(doubs) == "X"] <- "x"
names(doubs)[names(doubs) == "Y"] <- "y"
doubs$linear_pos <- build_linear_pos(doubs$dfs)

ggplot(doubs, aes(x, y, colour = dfs)) +
  geom_point(size = 2.3) +
  geom_path(alpha = 0.45) +
  labs(
    title = "Doubs sites (from DoubsSpa.csv)",
    subtitle = "Points connected in river-km order (dfs)",
    colour = "dfs"
  ) +
  theme_minimal(base_size = 12)
```

### Observed distance-decay from Doubs species data

```{r}
doubs_abund <- doubs_spe
names(doubs_abund)[1] <- "site"
site_coords_doubs <- doubs[, c("site", "x", "y", "linear_pos")]
attr(site_coords_doubs, "linear_wrap") <- FALSE

dd_doubs_eu <- calculate_distance_decay(doubs_abund, site_coords_doubs, metric = "euclidean")
dd_doubs_ap <- calculate_distance_decay(doubs_abund, site_coords_doubs, metric = "along_path")

dd_doubs <- rbind(
  transform(dd_doubs_eu, Metric = "Euclidean"),
  transform(dd_doubs_ap, Metric = "Along-path")
)

ggplot(dd_doubs, aes(Distance, Dissimilarity, colour = Metric)) +
  geom_point(alpha = 0.25) +
  geom_smooth(se = FALSE, method = "loess", span = 0.8, linewidth = 1.1) +
  labs(title = "Doubs observed distance-decay", y = "Bray-Curtis (binary)") +
  theme_minimal(base_size = 12)
```

## 2) Simulate in Doubs-like constrained network mode using real covariates

```{r}
P_doubs <- load_config(pkg_file("spesim_init_basic.txt", "examples"))
P_doubs$N_SPECIES <- 18
P_doubs$N_INDIVIDUALS <- 1300
P_doubs$DOMAIN_TYPE <- "network"
P_doubs$MODEL_FAMILY <- "hybrid"
P_doubs$NEUTRAL_M <- 0.10
P_doubs$DISPERSAL_KERNEL <- "exponential"
P_doubs$DISPERSAL_SCALE <- 0.9
P_doubs$DISPERSAL_DIRECTION_BIAS <- 0.25

P_doubs$SAMPLING_SCHEME <- "route"
P_doubs$ROUTE_QUADRAT_MODE <- "specified"
P_doubs$ROUTE_POSITIONS <- doubs$linear_pos
P_doubs$N_QUADRATS <- length(P_doubs$ROUTE_POSITIONS)
P_doubs$QUADRAT_SIZE_OPTION <- "small"

P_doubs$ENV_COVARIATES_FILE <- pkg_file("DoubsCovariates.csv", "extdata")
P_doubs$ENV_DRIVERS <- c("alt", "flo", "pH", "nit", "oxy")
P_doubs$GRADIENT_SPECIES <- c("A", "B", "C", "D", "E")
P_doubs$GRADIENT_ASSIGNMENTS <- c("alt", "flo", "pH", "nit", "oxy")
P_doubs$GRADIENT_OPTIMA <- c(alt = 0.7, flo = 0.4, pH = 0.5, nit = 0.3, oxy = 0.8)
P_doubs$GRADIENT_TOLERANCE <- c(alt = 0.25, flo = 0.30, pH = 0.25, nit = 0.25, oxy = 0.25)

P_doubs$INTERACTION_RADIUS <- 0
P_doubs$INTERACTIONS_FILE <- NULL
P_doubs$INTERACTIONS_EDGELIST <- NULL

dom_doubs <- make_domain_from_xy(doubs[, c("x", "y")], buffer_frac = 0.08)
res_doubs <- spesim_run(P_doubs, domain = dom_doubs, write_outputs = FALSE, seed = P_doubs$SEED, interactions_print = FALSE)

plot_spatial_sampling(
  res_doubs$domain, res_doubs$species_dist, res_doubs$quadrats, res_doubs$P,
  show_gradient = TRUE, env_gradients = res_doubs$env_gradients, gradient_type = "alt"
) + ggtitle("Simulated Doubs-like community using real Doubs covariates")
```

## 3) Seaweed coastline: real coordinates and covariates

```{r}
sea_spa <- read.csv(pkg_file("SeaweedSpa.csv", "extdata"), check.names = FALSE)
sea_env <- read_with_site(pkg_file("SeaweedEnv.csv", "extdata"))
sea_spe <- read_with_site(pkg_file("SeaweedSpe.csv", "extdata"))

sea <- transform(sea_spa, site = seq_len(nrow(sea_spa)), x = Longitude, y = Latitude)
sea <- merge(sea[, c("site", "x", "y")], sea_env, by = "site", sort = TRUE)
sea$linear_pos <- build_linear_pos(seq_len(nrow(sea)))

ggplot(sea, aes(x, y, colour = summerMean)) +
  geom_point(size = 2.1) +
  geom_path(alpha = 0.4) +
  labs(title = "Seaweed sections (real coordinates)", colour = "summerMean") +
  theme_minimal(base_size = 12)
```

### Observed distance-decay from seaweed species data

```{r}
sea_abund <- sea_spe
names(sea_abund)[1] <- "site"
site_coords_sea <- sea[, c("site", "x", "y", "linear_pos")]
attr(site_coords_sea, "linear_wrap") <- FALSE

dd_sea_eu <- calculate_distance_decay(sea_abund, site_coords_sea, metric = "euclidean")
dd_sea_ap <- calculate_distance_decay(sea_abund, site_coords_sea, metric = "along_path")

dd_sea <- rbind(
  transform(dd_sea_eu, Metric = "Euclidean"),
  transform(dd_sea_ap, Metric = "Along-path")
)

ggplot(dd_sea, aes(Distance, Dissimilarity, colour = Metric)) +
  geom_point(alpha = 0.15) +
  geom_smooth(se = FALSE, method = "loess", span = 0.8, linewidth = 1.1) +
  labs(title = "Seaweed observed distance-decay", y = "Bray-Curtis (binary)") +
  theme_minimal(base_size = 12)
```

### Simulate coastline-constrained community with seaweed covariates

```{r}
P_sea <- load_config(pkg_file("spesim_init_basic.txt", "examples"))
P_sea$N_SPECIES <- 26
P_sea$N_INDIVIDUALS <- 2200
P_sea$DOMAIN_TYPE <- "coastline"
P_sea$LINEAR_WRAP <- FALSE
P_sea$MODEL_FAMILY <- "hybrid"
P_sea$NEUTRAL_M <- 0.12
P_sea$DISPERSAL_KERNEL <- "gaussian"
P_sea$DISPERSAL_SCALE <- 0.2
P_sea$DISPERSAL_DIRECTION_BIAS <- 0.45

P_sea$SAMPLING_SCHEME <- "route"
P_sea$ROUTE_QUADRAT_MODE <- "specified"
P_sea$ROUTE_POSITIONS <- sea$linear_pos
P_sea$N_QUADRATS <- length(P_sea$ROUTE_POSITIONS)
P_sea$QUADRAT_SIZE_OPTION <- "small"

P_sea$ENV_COVARIATES_FILE <- pkg_file("SeaweedCovariates.csv", "extdata")
P_sea$ENV_DRIVERS <- c("summerMean", "winterMin", "augChl")
P_sea$GRADIENT_SPECIES <- c("A", "B", "C")
P_sea$GRADIENT_ASSIGNMENTS <- c("summerMean", "winterMin", "augChl")
P_sea$GRADIENT_OPTIMA <- c(summerMean = 0.65, winterMin = 0.55, augChl = 0.40)
P_sea$GRADIENT_TOLERANCE <- c(summerMean = 0.25, winterMin = 0.25, augChl = 0.30)

P_sea$INTERACTION_RADIUS <- 0
P_sea$INTERACTIONS_FILE <- NULL
P_sea$INTERACTIONS_EDGELIST <- NULL

dom_sea <- make_domain_from_xy(sea[, c("x", "y")], buffer_frac = 0.10)
res_sea <- spesim_run(P_sea, domain = dom_sea, write_outputs = FALSE, seed = P_sea$SEED, interactions_print = FALSE)

plot_spatial_sampling(
  res_sea$domain, res_sea$species_dist, res_sea$quadrats, res_sea$P,
  show_gradient = TRUE, env_gradients = res_sea$env_gradients, gradient_type = "summerMean"
) + ggtitle("Simulated coastline community using real seaweed covariates")
```

## Practical notes

- The CSV paths are wired with `pkg_file(...)`, which resolves
  `system.file("extdata/...", package = "spesim")` for installed-package use
  and falls back to `inst/extdata/...` for source-tree rendering.
- For Doubs, `X` and `Y` from `DoubsSpa.csv` are used as the site coordinates,
  and `dfs` from `DoubsEnv.csv` is used as the linear route coordinate.
- For seaweed, longitude/latitude are used as section coordinates, with an
  alongshore ordering converted to a linear coordinate.
- This provides real-world relevance now, while retaining internal synthetic
  generation for cases without external covariates.
