---
title: "Constrained ordination (db-RDA) on simulated spesim data"
output:
  rmarkdown::html_vignette:
    toc: true
    highlight: null
vignette: >
  %\VignetteIndexEntry{Constrained ordination (db-RDA) on simulated spesim data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>",
  warning = FALSE,
  message = FALSE,
  fig.width = 4.5,
  fig.height = 3.0,
  out.width = "85%",
  dpi = 300
)

ggplot2::theme_set(ggplot2::theme_bw(base_size = 8))
```

This vignette shows how to use **spesim** to generate a realistic (but known)
community dataset for practising constrained ordination.

We will:

1. Simulate a community in an irregular domain with **strong environmental gradients**.
2. Add **positive and negative interspecific interactions** (local neighbourhood effects).
3. Sample the landscape with quadrats to obtain:
   - a **site × species** abundance table
   - a **site × environment** table (mean environment per quadrat)
4. Fit a **distance-based RDA (db-RDA)** using **vegan**’s `capscale()`.
5. Create db-RDA plots in the same style as Tangled Bank’s *BCB743 constrained ordination* material.

## 1) Libraries

```{r}
library(spesim)
library(dplyr)
library(ggplot2)
library(vegan)
```

## 2) Configure a simulation with strong gradients + interactions

We start from the packaged basic init file and then modify it.

```{r}
`%||%` <- function(a, b) if (!is.null(a)) a else b

init_file <- system.file("examples/spesim_init_basic.txt", package = "spesim")
P <- load_config(init_file)

# --- keep the vignette fast-ish
P$N_INDIVIDUALS <- 3000L
P$N_SPECIES <- 16L
P$N_QUADRATS <- 40L
P$SAMPLING_RESOLUTION <- 70L
P$ENVIRONMENTAL_NOISE <- 0.02
P$ADVANCED_ANALYSIS <- FALSE

# --- strong filtering (narrow tolerances)
P$GRADIENT_SPECIES <- LETTERS[1:10]
P$GRADIENT_ASSIGNMENTS <- c(
  rep("temperature", 4),
  rep("elevation", 3),
  rep("rainfall", 3)
)
P$GRADIENT_OPTIMA <- 0.5
P$GRADIENT_TOLERANCE <- 0.06

# --- local interactions
# Values < 1 weaken neighbours (negative interactions); > 1 strengthen (positive).
P$INTERACTION_RADIUS <- 1.0
P$INTERACTIONS_EDGELIST <- c(
  # negative (competition/inhibition)
  "B,A,0.80",
  "C,A,0.85",
  "D,B,0.75",
  "E,D,0.80",

  # mild background pressure for some focal species (from all neighbours)
  "B,*,0.98",
  "C,*,0.98",

  # positive (facilitation)
  "F,G,1.25",
  "G,F,1.15",
  "H,I,1.20",
  "I,H,1.10"
)

# Reproducibility
P$SEED <- 2026L
```

Run the simulation:

```{r}
res <- spesim_run(P, write_outputs = FALSE, seed = P$SEED, quiet = TRUE)
```

## 3) The two tables we need for constrained ordination

### 3.1 Species table (responses)

`res$abund_matrix` is a site × species abundance table.

```{r}
A <- res$abund_matrix
head(A)
```

### 3.2 Environmental table (constraints)

`calculate_quadrat_environment()` provides mean environment per quadrat.

```{r}
E <- calculate_quadrat_environment(res$env_gradients, res$quadrats, sf::st_crs(res$domain))
head(E)
```

For constrained ordination we need the two tables aligned by site.

```{r}
dat <- dplyr::left_join(E, A, by = "site")
stopifnot(nrow(dat) == nrow(A))

# Response matrix (species): remove site column
Y <- dat |> dplyr::select(-site, -temperature_C, -elevation_m, -rainfall_mm)

# Constraint matrix (environment)
X <- dat |> dplyr::select(site, temperature_C, elevation_m, rainfall_mm)

# Standardise constraints (z-scores), as in Tangled Bank
Xz <- vegan::decostand(dplyr::select(X, -site), method = "standardize")
```

## 4) Fit db-RDA (vegan::capscale)

Distance-based RDA (db-RDA) is a constrained ordination that lets you choose a
non-Euclidean dissimilarity index. Here we use Bray–Curtis dissimilarity on the
abundance matrix.

```{r}
cap_full <- vegan::capscale(Y ~ temperature_C + elevation_m + rainfall_mm,
                           data = Xz,
                           distance = "bray",
                           add = TRUE)
cap_full
```

Permutation test for the overall constrained model:

```{r}
anova(cap_full)
```

Adjusted R² (variance explained by constraints):

```{r}
round(vegan::RsquareAdj(cap_full)$adj.r.squared, 3)
```

Check multicollinearity via variance inflation factors (VIF):

```{r}
vegan::vif.cca(cap_full)
```

## 5) Ordination plots (Tangled Bank style)

We create a clean ggplot of the first two constrained axes:

- **Sites** as points
- **Environmental vectors** as arrows
- Optional grouping by a simple “bioregion” proxy (temperature quartiles)

### 5.1 Extract scores

```{r}
sc_sites <- as.data.frame(vegan::scores(cap_full, display = "sites", scaling = 2))
sc_bp <- as.data.frame(vegan::scores(cap_full, display = "bp", scaling = 2))

sc_sites$site <- dat$site
sc_sites <- dplyr::left_join(sc_sites, X, by = c("site" = "site"))

# Simple temperature-based grouping (for teaching): 4 bins
sc_sites$temp_bin <- cut(sc_sites$temperature_C,
                         breaks = quantile(sc_sites$temperature_C, probs = seq(0, 1, 0.25), na.rm = TRUE),
                         include.lowest = TRUE)

# Arrow scaling: make biplot vectors visible in the same coordinate system
arrow_mul <- 1.3
sc_bp$var <- rownames(sc_bp)
sc_bp$CAP1 <- sc_bp$CAP1 * arrow_mul
sc_bp$CAP2 <- sc_bp$CAP2 * arrow_mul
```

### 5.2 Plot sites + biplot arrows

```{r}
cap_plot <- ggplot(sc_sites, aes(x = CAP1, y = CAP2)) +
  geom_hline(yintercept = 0, linewidth = 0.2, colour = "grey70") +
  geom_vline(xintercept = 0, linewidth = 0.2, colour = "grey70") +
  geom_point(aes(colour = temp_bin), size = 1.6, alpha = 0.85) +
  coord_equal() +
  labs(
    x = "db-RDA axis 1 (CAP1)",
    y = "db-RDA axis 2 (CAP2)",
    colour = "Temperature bin",
    title = "db-RDA on spesim quadrat data",
    subtitle = "Sites (quadrats) constrained by mean environment; Bray–Curtis distances"
  )

cap_plot
```

Add environmental vectors:

```{r}
cap_plot +
  geom_segment(
    data = sc_bp,
    aes(x = 0, y = 0, xend = CAP1, yend = CAP2),
    inherit.aes = FALSE,
    arrow = arrow(length = unit(0.15, "cm")),
    linewidth = 0.3,
    colour = "black"
  ) +
  geom_text(
    data = sc_bp,
    aes(x = CAP1, y = CAP2, label = var),
    inherit.aes = FALSE,
    size = 2.6,
    vjust = -0.6
  )
```

### 5.3 Sanity check: which gradients align with CAP1?

```{r}
sc_sites |>
  dplyr::summarise(
    cor_cap1_temp = cor(CAP1, temperature_C),
    cor_cap1_elev = cor(CAP1, elevation_m),
    cor_cap1_rain = cor(CAP1, rainfall_mm)
  )
```

## 6) Why this is useful for teaching

Because spesim generates:
- the **site × species** matrix (`abund_matrix`), and
- the corresponding **site × environment** matrix (via `calculate_quadrat_environment()`),

you can repeatedly create controlled datasets where you *know the truth* (strong
gradients and interactions were imposed) and then practise ordination workflows
that aim to recover the dominant drivers.

### Suggested exercises

- Increase/decrease `GRADIENT_TOLERANCE` and see how the constrained R² changes.
- Turn interactions off (`INTERACTION_RADIUS <- 0`) and compare db-RDA fits.
- Compare sampling schemes (random vs transect vs Voronoi) and see how sampling
  design alters ordination stability.
