---
title: "What is spesim? What isn't it?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{spesim model card (what it is, and what it isn't)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Purpose

**spesim** is designed for **teaching**, **methods testing**, and **exploratory
research**.

It helps you create a *known* spatial “truth” (individual locations, a sampling
design, and derived site \(\times\) species data) so you can:

- demonstrate concepts (environmental filtering, distance–decay, \(\beta\)-diversity),
- compare sampling schemes (random vs transect vs Voronoi, etc.),
- stress-test metrics and workflows on controlled scenarios.

## What the simulator generates

A typical run produces:

1. **A domain**: an `sf` polygon. The built-in domain is synthetic and uses
   arbitrary planar units.
2. **Environmental gradients**: gridded synthetic fields (temperature, elevation,
   rainfall) generated over the domain.
3. **A community of individuals**: `sf` point locations with species labels.
4. **Quadrat samples**: `sf` polygons representing the sampling design.
5. **Derived data**:
   - site \(\times\) species abundance matrix,
   - per-quadrat mean environment,
   - classic diagnostic summaries/plots (SAD, SAR, rarefaction, distance–decay).

## What the parameters mean

### Species–abundance distribution (SAD)

- Non-dominant species are drawn from a Fisher log-series tail.
- A configurable dominant fraction is allocated to a “dominant species”.

This is primarily a **pedagogical control**: it creates realistic-looking rank–
abundance curves without requiring a mechanistic population model.

### Environmental filtering

Species can be assigned to named gradients with an **optimum** and **tolerance**.

- “Optimum” = where along the gradient a species does best.
- “Tolerance” = how broad the response is.

This is implemented as a **Gaussian response** on a normalised 0–1 gradient
internally.

### Spatial structure and interactions

spesim supports several ways to introduce spatial structure:

- **CSR baseline** (Complete Spatial Randomness): homogeneous Poisson.
- **Clustering**: tunable clustering behaviour (including fast engines).
- **Inhibition**: “repulsion” style processes (Strauss/Geyer fast engines).
- **Neighbour effects**: a directed coefficient matrix applied within a single
  global interaction radius (values < 1 suppress; > 1 facilitate; 1 neutral).

These are intended to create **interpretable spatial patterns** for teaching and
experimentation.

## What spesim is *not*

spesim is **not** a full mechanistic ecological simulator. In particular, it does
*not* attempt to model:

- demography (births/deaths),
- explicit dispersal kernels as a population process,
- temporal dynamics or succession,
- detectability/observation error (unless you add it downstream),
- parameter inference / likelihood-based point process modelling.

If you need those, consider pairing spesim with specialised tools (e.g.
**spatstat** for point process inference/diagnostics, or domain-specific
individual-based/metacommunity simulators).

## Valid interpretations (and common misinterpretations)

**Valid:**

- “If I impose stronger environmental filtering, does distance–decay increase?”
- “How does a transect design compare to random quadrats for SAR shape?”
- “Do my diversity metrics behave sensibly under known gradients?”

**Not valid without extra work:**

- “This coefficient equals a real competition parameter.”
- “The ‘temperature’ units correspond to a real landscape unless I supplied one
  with a known CRS/units.”

## Reproducibility checklist

- Set (and record) `SEED` in your init file (or pass `seed=` to
  `spesim_run()`).
- If you regenerate advanced panels or vegan-derived summaries separately, call
  `set.seed(SEED)` immediately beforehand.

## Minimal example

```{r, eval=FALSE}
library(spesim)

P <- load_config(system.file("examples/spesim_init_basic.txt", package = "spesim"))
res <- spesim_run(P, write_outputs = FALSE, seed = 77)

plot_spatial_sampling(res$domain, res$species_dist, res$quadrats, res$P)
```
