---
title: "spesim recipes: environmental gradients"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{spesim recipes: environmental gradients}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

This short recipe shows how to turn **environmental filtering** on and off, and
how to change **optima/tolerances** for gradient-responsive species.

## Minimal run (no special gradients)

```{r}
library(spesim)

P <- load_config(system.file("examples/spesim_init_basic.txt", package = "spesim"))
P$N_SPECIES <- 10
P$N_INDIVIDUALS <- 1500
P$ADVANCED_ANALYSIS <- FALSE

res0 <- run_spatial_simulation(P = P, write_outputs = FALSE)
plot_spatial_sampling(res0$domain, res0$species_dist, res0$quadrats, res0$P)
```

## Make a few species respond strongly to a gradient

Here we keep everything else the same but make the temperature-responsive species
prefer warmer areas by shifting their **optimum** upward and tightening their
**tolerance**.

```{r}
P2 <- P

# A and B respond to temperature; C to elevation; D to rainfall (from the example init)
# Push temperature optimum to the warm end and make the response sharper
P2$GRADIENT_OPTIMA <- c(temperature = 0.80, elevation = 0.50, rainfall = 0.50)
P2$GRADIENT_TOLERANCE <- c(temperature = 0.07, elevation = 0.12, rainfall = 0.12)

res1 <- run_spatial_simulation(P = P2, write_outputs = FALSE)

p1 <- plot_spatial_sampling(res1$domain, res1$species_dist, res1$quadrats, res1$P)

# Show the underlying temperature field
p2 <- plot_spatial_sampling(
  res1$domain, res1$species_dist, res1$quadrats, res1$P,
  show_gradient = TRUE,
  env_gradients = res1$env_gradients,
  gradient_type = "temperature_C"
)

p1
p2
```

## Tip: reproducibility

If you want exact repeatability for gradient-driven patterns, set the seed (via
init file or `P$SEED`) and call `set.seed()` before running.
