---
title: "spesim: Doubs network method and interpolation workflow"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{spesim: Doubs network method and interpolation workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 5.5,
  warning = FALSE,
  message = FALSE
)
```

This vignette documents, in detail, the Doubs real-data workflow for:

1. constructing a directed site network from observed localities, and  
2. simulating species composition at unsampled positions between observed sites.

It uses only observed Doubs data from `inst/extdata`:

- `DoubsSpa.csv` (site coordinates),
- `DoubsEnv.csv` (site-level environment),
- `DoubsSpe.csv` (site x species abundances).

## Method summary

The current method is a linearized network MVP:

- Sites are ordered by an along-river coordinate (`dfs` in DoubsEnv).
- Edges are created between consecutive ordered sites.
- Edge weight is Euclidean segment length in projected site coordinates.
- Distance-decay along the network uses shortest-path distance over this edge graph.

Interpolation to unsampled sites uses:

- `linear_pos` between 0 and 1 as the 1D route coordinate,
- coordinate/environment interpolation along `linear_pos`,
- inverse-distance weighted (IDW-like) borrowing of observed community structure,
- optional directional bias to favor downstream or upstream influence.

## Load and join observed Doubs data

```{r}
library(spesim)
library(ggplot2)
library(sf)
library(dplyr)

pkg_file <- function(subpath, folder = "extdata") {
  p_inst <- system.file(folder, subpath, package = "spesim")
  if (nzchar(p_inst) && file.exists(p_inst)) return(p_inst)
  cand <- c(file.path("inst", folder, subpath), file.path("..", "inst", folder, subpath))
  for (p in cand) if (file.exists(p)) return(p)
  stop("Could not find: ", subpath)
}

read_with_site <- function(path) {
  x <- read.csv(path, check.names = FALSE)
  if ("" %in% names(x)) names(x)[names(x) == ""] <- "site"
  x
}

doubs_spa <- read_with_site(pkg_file("DoubsSpa.csv"))
doubs_env <- read_with_site(pkg_file("DoubsEnv.csv"))
doubs_spe <- read_with_site(pkg_file("DoubsSpe.csv"))

doubs <- merge(doubs_spa, doubs_env, by = "site", sort = TRUE)
names(doubs)[names(doubs) == "X"] <- "x"
names(doubs)[names(doubs) == "Y"] <- "y"
doubs$linear_pos <- (doubs$dfs - min(doubs$dfs)) / (max(doubs$dfs) - min(doubs$dfs))

site_coords_doubs <- doubs[, c("site", "x", "y", "linear_pos")]
site_coords_doubs$site <- as.character(site_coords_doubs$site)
doubs_abund <- doubs_spe
names(doubs_abund)[1] <- "site"
doubs_abund$site <- as.character(doubs_abund$site)
doubs_site_env <- doubs[, c("site", "dfs", "alt", "flo", "pH", "nit", "oxy", "bod")]
doubs_site_env$site <- as.character(doubs_site_env$site)
```

## Step 1: Create the network

### Construction rule

`build_network_edges()` creates a chain graph:

- sort sites by `linear_pos` (here derived from `dfs`),
- connect each site to the next one,
- compute edge weight as segment length between coordinates,
- if `directed = TRUE`, keep edge direction from low to high `linear_pos`.

```{r}
doubs_edges <- build_network_edges(
  site_coords = site_coords_doubs,
  order_by = "linear_pos",
  directed = TRUE
)

head(doubs_edges, 8)
```

### Plot the network geometry

```{r}
edge_xy <- doubs_edges |>
  left_join(site_coords_doubs |> select(site, x_from = x, y_from = y), by = c("from" = "site")) |>
  left_join(site_coords_doubs |> select(site, x_to = x, y_to = y), by = c("to" = "site"))

ggplot() +
  geom_path(
    data = site_coords_doubs[order(site_coords_doubs$linear_pos), ],
    aes(x, y),
    linewidth = 0.6,
    colour = "grey55",
    alpha = 0.6
  ) +
  geom_segment(
    data = edge_xy,
    aes(x = x_from, y = y_from, xend = x_to, yend = y_to, colour = weight),
    linewidth = 0.8,
    arrow = grid::arrow(length = grid::unit(1.6, "mm"), type = "closed")
  ) +
  geom_point(data = site_coords_doubs, aes(x, y), size = 2.2, colour = "black") +
  scale_colour_viridis_c(option = "D", end = 0.9) +
  labs(
    title = "Doubs directed site network",
    subtitle = "Edges connect consecutive sites in river-order (linear_pos from dfs)",
    colour = "Edge length"
  ) +
  theme_minimal(base_size = 12)
```

## Step 2: Build an observed-data `spesim_result` (no simulation yet)

```{r}
res_doubs_obs <- spesim_from_observed(
  abund_matrix = doubs_abund,
  site_coords = site_coords_doubs,
  site_env = doubs_site_env,
  edges = doubs_edges,
  directed = TRUE
)

res_doubs_obs
```

This object now supports the standard downstream analyses directly from real data:

- rank-abundance,
- occupancy-abundance,
- species-area accumulation,
- distance-decay,
- rarefaction.

## Step 3: Compare Euclidean vs network distance-decay on observed data

```{r}
dd_eu <- calculate_distance_decay(res_doubs_obs$abund_matrix, res_doubs_obs$site_coords, metric = "euclidean")
dd_net <- calculate_distance_decay(res_doubs_obs$abund_matrix, res_doubs_obs$site_coords, metric = "along_path")

dd <- rbind(
  transform(dd_eu, Metric = "Euclidean"),
  transform(dd_net, Metric = "Network shortest-path")
)

ggplot(dd, aes(Distance, Dissimilarity, colour = Metric)) +
  geom_point(alpha = 0.28) +
  geom_smooth(method = "loess", se = FALSE, span = 0.8, linewidth = 1.1) +
  labs(
    title = "Doubs observed distance-decay: Euclidean vs network distance",
    y = "Bray-Curtis (binary)"
  ) +
  theme_minimal(base_size = 12)
```

## Step 4: Simulate unsampled sites between observed sites

### Interpolation model used here

`simulate_between_observed()` does:

1. place new pseudo-sites along the `linear_pos` axis between min and max observed positions,
2. interpolate `x,y` and numeric environment along that axis,
3. estimate species probabilities from observed sites using inverse-distance weights in `linear_pos`,
4. optionally bias influence direction (`direction_bias > 0` favors downstream influence),
5. sample pseudo-site abundance vectors with multinomial draws.

```{r}
doubs_interp <- simulate_between_observed(
  res = res_doubs_obs,
  n_new_sites = 24,
  distance_power = 2,
  direction_bias = 0.35
)

head(doubs_interp$site_coords, 6)
head(doubs_interp$abund_matrix[, 1:8], 4)
```

### Plot observed + interpolated sites

```{r}
ggplot() +
  geom_point(data = res_doubs_obs$site_coords, aes(x, y), size = 2.1, colour = "black") +
  geom_point(data = doubs_interp$site_coords, aes(x, y), size = 1.8, colour = "red", alpha = 0.8) +
  labs(
    title = "Doubs observed (black) and interpolated unsampled sites (red)"
  ) +
  theme_minimal(base_size = 12)
```

## Step 5: Build a `spesim_result` for interpolated sites and analyse

```{r}
res_doubs_interp <- spesim_from_observed(
  abund_matrix = doubs_interp$abund_matrix,
  site_coords = doubs_interp$site_coords,
  site_env = doubs_interp$site_env,
  edges = build_network_edges(doubs_interp$site_coords, order_by = "linear_pos", directed = TRUE),
  directed = TRUE
)

res_doubs_interp
```

```{r, fig.width=12, fig.height=14}
panel_ready_res <- function(res, max_species = 26L) {
  spp <- setdiff(names(res$abund_matrix), "site")
  if (length(spp) <= max_species) return(res)
  sums <- colSums(res$abund_matrix[, spp, drop = FALSE], na.rm = TRUE)
  keep <- names(sort(sums, decreasing = TRUE))[seq_len(max_species)]
  out <- res
  out$abund_matrix <- cbind(site = res$abund_matrix$site, res$abund_matrix[, keep, drop = FALSE])
  out$species_dist <- res$species_dist[as.character(res$species_dist$species) %in% keep, ]
  out$P$N_SPECIES <- length(keep)
  out
}

generate_advanced_panel(panel_ready_res(res_doubs_obs)) +
  patchwork::plot_annotation(title = "Doubs observed data: advanced panel")
```

```{r, fig.width=12, fig.height=14}
generate_advanced_panel(panel_ready_res(res_doubs_interp)) +
  patchwork::plot_annotation(title = "Doubs interpolated unsampled sites: advanced panel")
```

## Notes and limitations

- This is still a linearized constrained-landscape approach.
- Network construction here is a chain over ordered observed sites, not yet a
  full branched river graph with explicit node-edge topology from hydrography.
- Interpolated unsampled communities are model-based pseudo-observations that
  preserve local observed structure but are not direct measurements.
- The workflow is meant to support method development and sensitivity analysis
  in realistic spatial constraints while full graph-theoretic routing is added.
